# What is a Venn diagram?

A Venn diagram is a deceptively simple figure,
intended to compare two or more sets in a visually intuitive way.

Each set is represented by a circle, drawn with a label beside it.
Two circles are shown below for sets `A` and `B`, and these circles
are drawn so they partially overlap.

The overlapping region represents the *intersection* of
sets `A` and `B`. Using mathematical symbols, the intersect is written
**$`r 'A∩B'`$**,
but we use `A&B` for convenience since the symbol **$`r '∩'`$**
does not exist on most computer keyboards.

```{r setup-01, include=FALSE}
suppressPackageStartupMessages(library(venndir))
suppressPackageStartupMessages(library(patchwork))
```

(ref:whatisvenn) A simple Venn diagram is shown with two sets: A and B.

```{r whatisvenn, echo=FALSE, fig.height=4, fig.width=7, out.width="65%", fig.cap="(ref:whatisvenn)", fig.alt="(ref:whatisvenn)"}
# setlist <- make_venn_test(n_sets=2, n_items=30,
#    set_names=c("A", "B"))
setlist <- counts2setlist(c(
   A=5, B=3, "A&B"=2))
venndir(setlist, poly_alpha=0, show_segments=FALSE, font_cex=2,
   innerborder="black", outerborder="black",
   draw_legend=FALSE,
   expand_fraction=c(-0.3, -0.3, -0.15, -0.3) + 0.2,
   outerborder.lwd=1, innerborder.lwd=1,
   legend_color_style=c("nofill", "blackborder"))
```

The numbers in the Venn diagram represent the number of items
in each logical grouping:

* '5' is the number of items in 'A' and 'not B'.
* '3' is the number of items in 'B' and 'not A'.
* '2' is the number of items in 'A' and 'B'.


## Problem Statement

> *"What problem are you trying to solve?" -- Lazar Arulnayagam*

When comparing biological effects on genes in two experiments,
two important questions arise:

1. Are the affected genes shared?
2. Are the changes in direction [concordant](#concordant)?

A standard Venn diagram can address shared gene membership,
but without directionality they fall short of answering question 2.

Venndir proposes a simple yet powerful extension to Venn diagrams:
placing [signed counts](#signed-counts) beside the [main counts](#main-counts).
The signed counts summarize directionality, indicating whether directions
of change were up or down. Direction is indicated using arrows and color cues,
red for changes up, and blue for changes down.

(ref:whatisvenndir) Venndir introduces an extension to a Venn diagram, with addition of directional counts beside each Venn overlap count. The direction is indicated by arrows up and down, and colors red and blue, respectively.

```{r whatisvenndir, echo=FALSE, fig.height=5, fig.width=9, out.width="100%", fig.cap="(ref:whatisvenndir)", fig.alt="(ref:whatisvenndir)"}
setlist_dir <- make_venn_test(n_sets=2,
   items=head(datasets::state.name, 20),
   sizes=c(5, 5),
   do_signed=TRUE,
   set_names=c("A", "B"))
venndir(setlist_dir,
   poly_alpha=0.4, fill=NA,
   show_segments=FALSE, font_cex=2,
   innerborder="black", outerborder="black",
   draw_legend=FALSE,
   expand_fraction=c(-0.1, -0.1, -0.05, -0.1),
   outerborder.lwd=1, innerborder.lwd=1,
   legend_color_style=c("nofill", "blackborder"))
```

For the simple example shown in Figure \@ref(fig:whatisvenndir),
the numbers are described as follows:

* There are **4** items in 'A' and 'not B'.

   * **2** of these items are UP.
   * **2** of these items are DOWN.

* There are **4** items in 'B' and 'not A'.

   * All **4** of these items are DOWN.

* There is **1** item shared by 'A' and 'B'.

   * This **1** item is UP.

Venndir also strives to answer *the very next question*:  
"What are those items?"

(ref:whatisvenndir-items) Venndir displays the items inside the diagram.

```{r whatisvenndir-items, echo=FALSE, fig.height=5, fig.width=9, out.width="100%", fig.cap="(ref:whatisvenndir-items)", fig.alt="(ref:whatisvenndir-items)"}
vi <- venndir(setlist_dir,
   xyratio=1.8, item_buffer=0, width_buffer=0.05,
   show_labels="Ni",
   item_cex=c(2, 2, 2.2) * 0.8 * 1.1,
   poly_alpha=0.4, fill=NA,
   show_segments=FALSE, font_cex=2,
   innerborder="black", outerborder="black",
   draw_legend=FALSE,
   expand_fraction=c(-0.1, -0.1, -0.05, -0.1),
   outerborder.lwd=1, innerborder.lwd=1)
```


```{r todo-intro, echo=FALSE, eval=FALSE}
# When to use a Venn diagram

# How to interpret a Venn diagram

# How to interpret directional Venn diagrams

# Guidance for directional Venn diagrams

# Guidance for Venndir
```
