# What is a Venn diagram?

A venn diagram is a deceptively simple figure,
intended to compare two or more sets in a visually intuitive way.

Each set is represented by a circle, drawn with a label beside it.
Two circles are shown below for sets `A` and `B`, and these circles
are drawn so they partially overlap.

The overlapping region represents the *intersection* of
sets `A` and `B`. Using mathematical symbols, the intersect is written
**$`r 'A∩B'`$**,
but we use `A&B` for convenience since the symbol **$`r '∩'`$**
does not exist on most computer keyboards.

(ref:whatisvenn) A simple Venn diagram is shown with two sets: A and B.

```{r whatisvenn, echo=FALSE, fig.height=4, fig.width=7, out.width="65%", fig.cap="(ref:whatisvenn)", fig.alt="(ref:whatisvenn)"}
# setlist <- make_venn_test(n_sets=2, n_items=30,
#    set_names=c("A", "B"))
setlist <- counts2setlist(c(
   A=5, B=3, "A&B"=2))
venndir(setlist, poly_alpha=0, show_segments=FALSE, font_cex=2,
   innerborder="black", outerborder="black",
   draw_legend=FALSE,
   expand_fraction=c(-0.3, -0.3, -0.15, -0.3),
   outerborder.lwd=1, innerborder.lwd=1,
   legend_color_style=c("nofill", "blackborder"))
```

The numbers in the Venn diagram represent the number of items
in each logical grouping:

* '5' is the number of items in 'A' and 'not B'.
* '3' is the number of items in 'B' and 'not A'.
* '2' is the number of items in 'A' and 'B'.


## Problem Statement

> *"What problem are you trying to solve?" -- Lazar Arulnayagam*

When comparing biological effects on genes in two experiments,
two important questions arise:

1. Are the genes shared? **(Venn)**
2. Are the changes [concordant](#concordant) in direction? **(Venndir)**

A Venn diagram does not include direction, and cannot answer question 2.

Venndir proposes a simple extension to a Venn diagram, the placement
of [signed counts](#signed-counts) beside the [main counts](#main-counts).
The [signed counts](#signed-counts) summarize
the directionality of the [main counts](#main-counts), and are described using
directional arrows, up and down, and are reinforced using color,
red for up, and blue for down.

(ref:whatisvenndir) Venndir introduces an extension to a Venn diagram, with addition of directional counts beside each Venn overlap count. The direction is indicated by arrows up and down, and colors red and blue, respectively.

```{r whatisvenndir, echo=FALSE, fig.height=4, fig.width=7, out.width="65%", fig.cap="(ref:whatisvenndir)", fig.alt="(ref:whatisvenndir)"}
setlist_dir <- make_venn_test(n_sets=2, n_items=30,
   sizes=c(7, 5),
   do_signed=TRUE,
   set_names=c("A", "B"))
venndir(setlist_dir,
   poly_alpha=0.4, fill=NA,
   show_segments=FALSE, font_cex=2,
   innerborder="black", outerborder="black",
   draw_legend=FALSE,
   expand_fraction=c(-0.3, -0.3, -0.15, -0.3),
   outerborder.lwd=1, innerborder.lwd=1,
   legend_color_style=c("nofill", "blackborder"))
```


```{r 01-todo, eval=FALSE, include=FALSE}
## Todo

# * Introduce other R packages used for detailed customization
#    * Venndir makes extensive use of grid graphics, including gridGeometry
#    * Venndir relies heavily upon marquee, and thus systemfonts by proxy, for font handling

```
