# Data Import

```{r setup-02, include=FALSE}
knitr::opts_chunk$set(
   # fig.retina = 2, out.width = '100%'
   cache=TRUE,
   fig.align="center",
   fig.width=6,
   fig.height=5
)
```

```{r setup-02b, include=FALSE}
# from
# https://blog.datascienceheroes.com/how-to-self-publish-a-book-customizing-bookdown/
# https://blog.datascienceheroes.com/how-to-self-publish-a-book/
# knitr::opts_chunk$set(fig.pos='H')
```

The first key step in using Venndir is to import data in a
recognized format. There are several formats available:

* **[List of sets]**, referred to as a `setlist`
   * a `list` of `character` item vectors.
   * sets are represented by `names(setlist)`
   
* **[List of signed sets]**, referred to as a signed `setlist`
   * a `list` of *signed* `numeric` vectors.
   * signed vectors use values `-1` or `1`, and store items in `names()`.
   * sets are represented by `names(setlist)`

* **[Incidence matrix]**
   * a `numeric` matrix with items as `rownames()`, sets as `colnames()`.
   * items are indicated by the value `1` in the appropriate
   row and column.

* **[Signed incidence matrix]**
   * a `numeric` matrix with items as `rownames()`, sets as `colnames()`.
   * items are indicated by `-1` or `1` in the appropriate
   row and column.

* **[Overlap counts]**
   * a `numeric` vector with the number of items in each overlap.
   * the overlap name is stored in `names()`, in format: `'A'`, `'B'`,
   `'A&B'`.

* **[Signed overlap counts]**
   * a `list` of `numeric` vectors, named by overlap set.
   * each overlap is also a `list` named by the directionality.

* **Other formats**
   * [Items]
   * [Overlap list]
   * [Signed overlap list]
   * [List of data frames]

* **Specialized formats**
   * [limma]
   * [DESeq2]

## Common Data Input Types

### List of sets

The simplest input is a List of [sets](#set), where each set is
represented as an R `vector` of items.

```{r input-list}
setlist <- make_venn_test()
str(setlist)
```

Each vector is not considered 'signed', because each vector has no `names()`.

This `setlist` can be used directly with `venndir()`.
In Figure \@ref(fig:input-list-venndir) there are no
[signed counts](#signed-counts) displayed.

(ref:input-list-venndir) Default Venn diagram with three sets.

```{r input-list-venndir, fig.height=6, fig.cap="(ref:input-list-venndir)", fig.alt="(ref:input-list-venndir)"}
venndir(setlist)
```

### List of signed sets

The next common input is a list of [signed sets](#signed-set),
represented as `numeric` vectors with items stored as `names()`.

An example is shown below, showing only the first six elements of each
vector. Notice the vector names are `"item_014"`, `"item_195"`,
while the values are `-1`, `-1`.

```{r input-signed-list}
setlist <- make_venn_test(do_signed=TRUE)
lapply(setlist, head, 6)
```

Figure \@ref(fig:input-signed-list-venndir) illustrates the
output from `venndir()`, where
[signed counts](#signed-counts) are displayed using
`overlap_type='concordance'` by default for a
[signed setlist](#signed-setlist).

(ref:input-signed-list-venndir) Default Venn diagram with three signed sets. The overlap counts are each tabulated by directional sign.

```{r input-signed-list-venndir, fig.cap="(ref:input-signed-list-venndir)", fig.alt="(ref:input-signed-list-venndir)"}
venndir(setlist)
```

The output is described below:

* The region `"set_A"` contains `19` items that are not present in any
other set for this Venn diagram.
   * `10` of these `19` items are "up", indicated by the red label
   which includes the up arrow $`r "\u2191 10"`$.
   * `9` of these `19` items are "down", indicated by the blue label
   which includes the down arrow $`r "\u2193 9"`$.

* The region of overlap between `'set_A'` and `'set_B'` contains 5 items
which are only present in these two sets, and not present in `'set_C'`.
   * `3` of these `5` items are "up" in both `'set_A'` and `'set_B'`,
   indicated by the red label with two up arrows: $`r "\u2191\u2191 3"`$.
   * `2` of these `5` items are discordant in direction, which means
   the sign in `'set_A'` disagrees with `'set_B'`. The label is indicated
   in grey color and large `'X'` $`r "X 2"`$.
   The purpose of using one `'X'` is to avoid describing all possible
   combinations of "up" and "down".

The default summary for a [signed setlist](#signed-setlist) is
`overlap_type="concordance"`,
which summarizes directional discordance in one category `'X'`, and
tabulates counts in each subset that involves only one direction.

For other approaches to summarize directional counts,
see the summary in Table \@ref(tab:overlap-type-table), in
the section [Overlap Type].


### Incidence matrix

An [incidence matrix](#g-incidence-matrix) (im) is another common input format,
a `matrix` whose `rownames` are items, and `colnames` are sets.
Any non-zero, non-empty value in the matrix indicates the item (row) exists
in the set (column).

```{r input-im}
setlist <- make_venn_test()
im <- list2im_opt(setlist)
head(im, 10)
```

Figure \@ref(fig:input-im-venndir) shows the Venn diagram
created by `venndir()` accepts an incidence matrix as input data.
The input data is converted to a `setlist` within the function.

(ref:input-im-venndir) Default Venn diagram with three sets, using an incidence matrix as input data.

```{r input-im-venndir, fig.cap="(ref:input-im-venndir)", fig.alt="(ref:input-im-venndir)"}
venndir(im)
```

When the [incidence matrix](#g-incidence-matrix) only contains positive values, it is
assumed to be non-directional. This assumption can be changed
by using the argument `overlap_type`.

### Signed incidence matrix

A [signed incidence matrix](#g-signed-incidence-matrix) is similar to an incidence matrix,
except that it may include positive and negative values.

```{r input-signed-im}
setlist <- make_venn_test(do_signed=TRUE)
im <- list2im_value(setlist)
head(im, 10)
```

Figure \@ref(fig:input-signed-im-venndir) shows the signed Venn
diagram created by `venndir()` also accepts a
[signed incidence matrix](#signed-incidence-matrix)
as input data, and subsequently displays signed counts.

(ref:input-signed-im-venndir) Default Venn diagram with three sets, using a signed incidence matrix as input data.

```{r input-signed-im-venndir, fig.height=6, fig.cap="(ref:input-signed-im-venndir)", fig.alt="(ref:input-signed-im-venndir)"}
venndir(im)
```


### Overlap counts

If the [Venn overlap](#venn-overlap) counts are already known, they can also
be used to re-create the corresponding Venn diagram.

The function `counts2setlist()` accepts Venn counts, and returns
the corresponding `setlist`.
The counts should be named by the [Venn overlap](#venn-overlap), by using the name
of each set involved, separated by ampersand `'&'`.
The example below uses sets `"A"` and `"B"`, and the corresponding
overlap between `'A'` and `'B'` is named `'A&B'`.

Note that the overlap name should be defined in quotes in R code,
so the ampersand `'&'` is stored properly.

```{r input-counts}
counts <- c(
   "A"=12,
   "B"=9,
   "B&A"=15)
setlist <- counts2setlist(counts)
str(setlist)
```

Figure \@ref(fig:input-counts-venndir) shows the output of `venndir()`
using this `setlist` as input.

(ref:input-counts-venndir) Venn diagram created by using overlap counts as input data.

```{r input-counts-venndir, fig.cap="(ref:input-counts-venndir)", fig.alt="(ref:input-counts-venndir)"}
venndir(setlist)
```

:::: {.tipbox data-latex=""}

**Tip:**

When starting with overlap counts, it is recommended to define
set names with single characters, such as `'A'`, `'B'`, and `'C'`.

* The set names can be adjusted afterwards by editing `names(setlist)`.
* However, the preferred approach is to use
`venndir()` arguments `setlist_labels` and `legend_labels`,
also described in [Custom legend labels].
* Figure \@ref(fig:input-counts-venndir-2) illustrates this process.

The complete combination of sets and overlaps can be defined by
calling `make_venn_combn_df()` using a vector of set names.  
For example `rownames(make_venn_combn_df(LETTERS[1:3]))` produces
the following:  
`r jamba::cPaste(sep=", ", paste0("'", rownames(make_venn_combn_df(LETTERS[1:3])), "'"))`

::::

(ref:input-counts-venndir-2) Venn diagram using custom set names using `setlist_labels` and `legend_labels`.

```{r input-counts-venndir-2, fig.cap="(ref:input-counts-venndir-2)", fig.alt="(ref:input-counts-venndir-2)"}
venndir(setlist,
   setlist_labels=c("Set A:\ncontrol state",
      "Set B:\ntest state"),
   legend_labels=c("A: control state",
      "B: test state"))
```


### Signed overlap counts

Similar to providing overlap counts as above, this approach defines
counts for each directional overlap, using `signed_counts2setlist()`.
This input is quite complex, also the least common.

The input is a `list` of `integer` count vectors. Each vector is
named by the overlap, for example `"A"` or `"A&B"`. Each vector
is named by the direction, delimited by underscore `"_"`, for
example `"1"` for 'up', or `"1_-1"` for 'up_down'.

Notice the format of the input data:

```{r input-signed-counts}
signed_counts <- list(
   "A"=c(
      "1"=80,
      "-1"=95),
   "B"=c(
      "1"=15,
      "-1"=30),
   "A&B"=c(
      "1_1"=100,
      "1_-1"=3,
      "-1_1"=4,
      "-1_-1"=125))
signed_counts
```

This input format is complicated for 2-way data, and certainly
even more complicated for 3-way data. However, sometimes it
is the most practical way to produce a given figure.

Counts are converted to a signed setlist with `signed_counts2setlist()`.
Item names are generated only to create the `setlist` and
are not otherwise useful.

```{r input-signed-counts2}
setlist <- signed_counts2setlist(signed_counts)
lapply(setlist, jamba::middle, 5)
```

Figure \@ref(fig:input-signed-counts-venndir) shows the `setlist`
visualized with `venndir()`.

(ref:input-signed-counts-venndir) Venndir showing signed counts summarized using `overlap_type="each"`, which includes counts for each combination of signs.

```{r input-signed-counts-venndir, fig.height=6, fig.width=5, fig.cap="(ref:input-signed-counts-venndir)", fig.alt="(ref:input-signed-counts-venndir)"}
venndir(setlist, overlap_type="each")
```


### Overlap list

This approach is conceptually similar to [Overlap counts]
which starts with the Venn counts as input, except in this case
instead of supplying the integer count in each [Venn overlap](#venn-overlap),
the data contains the actual items.

Consider a simple example:

* two sets: 'A' and 'B'
* three total items, one only present in 'A', one unique to 'B', and
one shared by both groups which is assigned to 'A&B'.

```{r overlaplist-1}
overlaplist <- list(
   A=c("Christina"),
   B=c("James"),
   "A&B"=c("Jillian", "Zander", "Java Pup")
)
str(overlaplist)
```


This `overlaplist` is converted into `setlist` using `overlaplist2setlist()`,
which can be input to `venndir()`.

Figure \@ref(fig:overlaplist-2) shows the default Venn diagram (left),
and a variation which displays the item labels (right) using argument
`show_labels="Ni"`.
The `show_labels` argument is described in [Label Content].

(ref:overlaplist-2) Venn diagram after converting an overlap list to `setlist`, showing overlap counts (left), and overlap item labels (right).

```{r overlaplist-2, fig.ncol=2, fig.height=5, fig.width=4, out.width="50%", fig.cap="(ref:overlaplist-2)", fig.alt="(ref:overlaplist-2)", fig.subcap=c("Overlap counts.", "Overlap item labels.")}
setlist <- overlaplist2setlist(overlaplist)
venndir(setlist)

venndir(setlist,keep_item_order=TRUE,
   item_cex_factor=c(0.5, 0.5, 0.8),
   show_labels="Ni")
```

The argument `item_cex_factor` is used to adjust the item label font
size, and is described in [Item Labels].


### Signed overlap list

A powerful but complex import option is a signed overlap list,
similar to the previous section [Overlap list]
with the addition of directional sign.

The format is similar to overlap list:

* `list` named by the [Venn overlap](#venn-overlap): 'A', 'B', 'A&B'
* Each list element is also a `list`, named by the directional sign.

   * Each sign is defined using `'1'` or `'-1'` separated by a space.
   * Signs involving one set: `"1"`, `"-1"`
   * Signs involving two sets: `"1 1"`, `"1 -1"`, etc.
   * Signs involving three sets: `"1 1 1"`, `"1 -1 1"`, etc.

```{r signed-overlaplist-1}
signed_overlaplist <- list(
   A=list(
      "-1"=c("Item_A1"),
      "1"=c("Item_A2")),
   B=list(
      "1"=c("Item_B1", "ItemB2"),
      "-1"=c("ItemB3")),
   "A&B"=list(
      "1 1"=c("Item_AB2"),
      "1 -1"=c("Item_AB4"),
      "-1 1"=c("Item_AB1"),
      "-1 -1"=c("Item_AB3"))
)
str(signed_overlaplist)
```

The `signed_overlaplist` is converted to `setlist` using
`overlaplist2setlist()`.

Figure \@ref(fig:signed-overlaplist-2) shows the resulting Venn
diagram with item labels enabled. Notice the item labels include
a directional arrow, and are colored by the sign.

(ref:signed-overlaplist-2) Venn diagram derived from a signed overlap list. The figure shows item labels which are colored by sign, and placed beside a directional arrow.

```{r signed-overlaplist-2, fig.cap="(ref:signed-overlaplist-2)", fig.alt="(ref:signed-overlaplist-2)"}
setlist <- overlaplist2setlist(signed_overlaplist)
v <- venndir(setlist,
   show_labels="Ni",
   item_cex_factor=0.6,
   xyratio=1.5)
```

This input format can also be generated from the `Venndir` object
itself using `overlaplist(v)`. However, it requires that `venndir()`
was called with argument `overlap_type="each"` to preserve each sign.
The default `overlap_type="concordance"` does not preserve signs
for [discordant](#discordant) overlaps.


#### Alternative signed overlap list

An alternative input format is shown below, which may be more convenient
to produce in some circumstances.

The alternate format is shown below:

```{r alt-signed-overlaplist-1}
alt_signed_overlaplist <- list(
   A=c("Item_A1"="-1",
      "item_A2"="1"),
   B=c("Item_B1"="1",
      "ItemB2"="1",
      "ItemB3"="-1"),
   "A&B"=c("Item_AB2"="1 1",
      "Item_AB4"="1 -1",
      "Item_AB1"="-1 1",
      "Item_AB3"="-1 -1"))
alt_signed_overlaplist
```

First, the alternative format is converted to the expected
format for `overlaplist2setlist()`.
Then it is converted to `setlist`.

```{r alt-signed-overlaplist-2-prep}
signed_overlaplist <- lapply(alt_signed_overlaplist, function(i){
   split(names(i), i)
})
setlist <- overlaplist2setlist(signed_overlaplist)
```

The resulting `setlist` can be used with `venndir()` as
in Figure \@ref(fig:signed-overlaplist-2).


### Items

A streamlined approach to convert items from a `vector` or `list`
into `setlist` is provided by `venn_meme()`, as described [Venn Memes].

The process assumes the input data is provided in a specific order.
The process is quite similar to that described in [Overlap list],
except that this process assumes the order matches the output of
`make_venn_combn_df()`.

The assumptions:

* Input must be provided in the order defined by
`make_venn_combn_df(n)` where `n` is the number of sets, or
`make_venn_combn_df(x)` where `x` is a vector of set names.
* Input is provided without using overlap names, they are assigned
based upon the length of the input data.
* Duplicate items are not permitted.

To illustrate the process, each example below uses a vector of
character `LETTERS`, showing how the letters are populated into
the Venn diagram.

Figure \@ref(fig:import-meme-all) illustrates the process by showing the
output of `venn_meme()`, with each panel using a slightly longer
vector of items.

* For 1 item, the output includes one set (not shown).
* For 2 or 3 items, it produces two sets (a).
* For 3 to 7 items, a 3-way Venn diagram is created (b)
* For 8 to 15 items, a 4-way Venn diagram is created (c).
* For 16 to 31 items, a 5-way Venn diagram is created (d).

```{r import-meme-n3, eval=FALSE}
vm3 <- venn_meme(head(LETTERS, 3))
vm7 <- venn_meme(head(LETTERS, 7))
vm15 <- venn_meme(head(LETTERS, 15))
vm31 <- venn_meme(1:31)
```

(ref:import-meme-all) Four Venn diagrams showing the output from `venn_meme()` when providing 3, 7, 15, and 31 items, respectively.

```{r import-meme-all, echo=FALSE, fig.height=8, fig.width=9, out.width="90%", fig.cap="(ref:import-meme-all)", fig.alt="(ref:import-meme-all)"}
vm3 <- venn_meme(head(LETTERS, 3), do_plot=FALSE, font_cex=0.5, item_cex_factor=0.5)
vm7 <- venn_meme(head(LETTERS, 7), do_plot=FALSE, font_cex=0.5, item_cex_factor=0.5)
vm15 <- venn_meme(head(LETTERS, 15), do_plot=FALSE, font_cex=0.5, item_cex_factor=0.5)
vm31 <- venn_meme(1:31,
   do_plot=FALSE, font_cex=0.5, item_cex_factor=0.5)

use_expand <- -1 * c(0.1, 0.3, 0.1, 0.3);

vm3_gt <- attr(plot(vm3, # main="head(LETTERS, 3)",
   expand_fraction=c(1,1,1,1)*use_expand*2,
   do_draw=FALSE), "gtree")
vm7_gt <- attr(plot(vm7, # main="head(LETTERS, 7)",
   expand_fraction=c(1,1,1,1)*use_expand*1.5,
   do_draw=FALSE), "gtree")
vm15_gt <- attr(plot(vm15, # main="head(LETTERS, 15)",
   expand_fraction=c(1,2,3.5,2)*use_expand*1.0,
   do_draw=FALSE), "gtree")
vm31_gt <- attr(plot(vm31, # main="head(LETTERS, 31)",
   expand_fraction=c(1,1,1,1)*use_expand,
   do_draw=FALSE), "gtree")

vm4 <- (wrap_elements(vm3_gt) +
   wrap_elements(vm7_gt)) /
   (wrap_elements(vm15_gt) +
   wrap_elements(vm31_gt)) +
   plot_layout(widths=c(1, 1)) +
   plot_annotation(tag_levels='a')
vm4

```


Tip: A useful by-product of calling `venn_meme()` is the creation
of a `setlist` or `overlaplist`, without needing to provide
all the [Venn overlap](#venn-overlap) labels upfront.

```{r import-meme-n3-setlist}
vm3 <- venn_meme(head(LETTERS, 3), do_plot=FALSE)
setlist(vm3)
overlaplist(vm3)
```


### List of data frames

A common starting point is a `data.frame` for each set,
with a column of numeric values, and column of item names.
Very often this input is a "table of stats" which may
also need to be filtered for statistical significance.

The general guidance is to filter each `data.frame`, then convert
the result into a vector. The vector can either be un-signed,
or signed.

1. **[Setlist](#setlist)**: a `character` vector of items.
2. **[Signed setlist](#signed-setlist)**: vector using `sign()`, named by item.

To illustrate, consider the test data generated below, as two
`data.frame` objects with P-values and log2 fold changes.

```{r input-df-list}
# define some test data
set.seed(123)
df1 <- data.frame(
   item=paste0("item", jamba::padInteger(1:20)),
   pvalue=round(digits=3, stats::runif(20) / 5),
   log2FoldChange=round(digits=3, rnorm(20) * 3))

df2 <- data.frame(
   item=paste0("item", jamba::padInteger(1:20)),
   pvalue=round(digits=3, stats::runif(20) / 5),
   log2FoldChange=round(digits=3, rnorm(20) * 3))

head(df1)
```

#### Option 1. Setlist {-}

One convenient approach is to create a `list` of `data.frame` objects,
apply a P-value filter to each entry, and return column `'item'`.

> This output does not have directional sign.

```{r input-df-list-1, fig.height=5, fig.width=8}
# create a list of data.frame objects
dflist <- list(Dataset1=df1,
   Dataset2=df2)

# iterate the list
setlist <- lapply(dflist, function(df){
   # apply stat threshold
   dfsub <- subset(df, pvalue < 0.1);
   # return the item name
   dfsub$item
})
str(setlist)
```

#### Option 2. Signed setlist {-}

The other common alternative is to extend Option 1 to create
a signed list, using the sign of `'log2FoldChange'`.
The vector should be the sign, then *named* by 'item'.

```{r input-df-signed-list, fig.height=5, fig.width=8}
# create a list of data.frame objects
dflist <- list(Dataset1=df1,
   Dataset2=df2)

# iterate the list
setlist_signed <- lapply(dflist, function(df){
   # apply stat threshold
   dfsub <- subset(df, pvalue < 0.1);
   # create vector using sign()
   x <- sign(dfsub$log2FoldChange);
   # add item names
   names(x) <- dfsub$item;
   x
})
setlist_signed
```

Figure \@ref(fig:input-df-signed-list-venndir) shows the resulting Venn diagram without sign (left) and with sign (right).

(ref:input-df-signed-list-venndir) Venn diagram using a list of data frames to a `setlist` (left) and signed `setlist` (right).

```{r input-df-signed-list-venndir, echo=FALSE, fig.ncol=2, fig.height=5, fig.width=7, out.width="50%", fig.cap="(ref:input-df-signed-list-venndir)", fig.alt="(ref:input-df-signed-list-venndir)", fig.subcap=c("setlist", "signed setlist")}
venndir(setlist, font_cex=1.5)
venndir(setlist_signed, font_cex=1.5)
```

```{r input-df-signed-list-venndir-1, eval=FALSE}
venndir(setlist)
venndir(setlist_signed)
```


## Specialized Input Types

The driving motivation for Venndir involved gene expression data,
and there are several common methods to import differentially
expressed genes.

### limma

The **limma** [@R-limma] package covers an amazing array of gene and
Omics expression analysis methods, spanning transcript and proteomics
microarray data, transcriptomics RNA sequencing (RNA-seq) data, and
a broad range of other data types including proteomics, lipidomics,
metabolomics, and DNA methylation.

The common paradigm, beginning with `x` as expression data.
See `limma::getEAWP()` for a description of all recognized data
object types.

```{r limma-example-1, eval=FALSE}
library(limma)

# fit linear model
fit <- lmFit(x, design=design)

# apply contrasts
fit2 <- contrasts.fit(fit, contrasts=contrasts)

# Optional Empirical Bayes moderated t-statistic
fit3 <- eBayes(fit)
```

Now one would use `fit3`, or `fit2` if eBayes was not necessary, to
summarize the statistical contrasts.

Two common approaches are used here, both can be valid:

1. Extract each contrast into a `data.frame` using `limma::topTable()`,
then follow [List of data frames].
2. Use `limma::decideTests()` to filter all contrasts together.
This function returns `limma::TestResults-class` which is equivalent to
[Signed incidence matrix].

For example, to filter by adjusted P-value 0.05, and 1.5-fold change:

```{r limma-example-2, eval=FALSE}
im <- limma::decideTests(fit3,
   p.value=0.05,
   lfc=log2(1.5))
```

The resulting [incidence matrix](#incidence-matrix) `im` contains one
column per contrast.

When using only a subset of contrasts, use the argument `sets` as
described in [Consistent Set Colors].
The example below chooses `sets=c(1, 3)`.

```{r limma-example-3, eval=FALSE}
venndir(im, sets=c(1, 3))
```

### DESeq2

**DESeq2** [@R-DESeq2] is widely used and excellent for RNA-seq analysis.
The example below is adapted from `DESeq2::DESeq()` to use three groups:
`'A'`, `'B'`, `'C'`.

```{r deseq2-example1}
suppressPackageStartupMessages(library(DESeq2))
set.seed(123)
cnts <- matrix(rnbinom(n=1500, mu=100, size=1/0.5), ncol=15)
rownames(cnts) <- paste0("row", jamba::padInteger(1:100))
cond <- factor(rep(LETTERS[1:3], each=5))
colnames(cnts) <- jamba::makeNames(cond, suffix="")
cnts[1, 1:5 + 0] <- cnts[1, 1:5] + 275;
cnts[2, 1:5 + 5] <- cnts[2, 1:5 + 5] + 355;
cnts[3, 1:5 + 10] <- cnts[3, 1:5 + 10] + 425;
cnts[4, 1:5 + 10] <- cnts[4, 1:5 + 10] + 655;

# object construction
suppressMessages(
   dds <- DESeqDataSetFromMatrix(cnts, DataFrame(cond), ~ cond)
)

# standard analysis
suppressMessages(
   dds <- DESeq(dds)
)
```

The object `dds` can be used to extract statistical results
for each contrast, for example see `DESeq2::resultsNames(dds)`.

The next steps extract a `DESeqResults` object for each contrast
of interest. Note: Use whichever method you typically use at
this point, there are too many possible workflows to cover here.


```{r deseq2-example2, eval=FALSE}
# Contrast: 'B - A'
resBA <- results(dds, contrast=c("cond", "B", "A"))

# Contrast: 'C - A'
resCA <- results(dds, contrast=c("cond", "C", "A"))

# Print the first few rows of the "C - A" contrast results
head(data.frame(check.names=FALSE, resCA))
```

```{r deseq2-example2-real}
# Contrast: 'B - A'
resBA <- results(dds, contrast=c("cond", "B", "A"))

# Contrast: 'C - A'
resCA <- results(dds, contrast=c("cond", "C", "A"))

# Print the first few rows of the "C - A" contrast results
idf <- head(data.frame(check.names=FALSE, resCA))
idf$baseMean <- round(idf$baseMean, digits=1)
idf$log2FoldChange <- round(idf$log2FoldChange, digits=2)
idf$lfcSE <- round(idf$lfcSE, digits=2)
idf$stat <- round(idf$stat, digits=2)
kdf <- knitr::kable(
   booktabs=TRUE,
   idf,
   caption="The first 6 rows following DESeq2 statistical analysis.") |>
   kableExtra::column_spec(1, bold=TRUE)
kdf

```

Two steps are applied, intentionally lenient for this example:

1. Filter by statistical thresholds
   * adjusted P-value `padj < 0.20`
   * fold change `abs(log2FoldChange) > log2(1)`

2. Convert to [signed set](#signed-set), assigning names using the DESeq2
result rownames.

```{r deseq2-example3-ba}
# filter statistics
resBA_hitdf <- subset(resBA, 
   padj <= 0.2 & 
   abs(log2FoldChange) > log2(1))

# generate named vector
resBA_hits <- setNames(
   sign(resBA_hitdf$log2FoldChange),
   rownames(resBA_hitdf))

resBA_hits
```

The same steps are performed for contrast 'C - A'.

```{r deseq2-example3-ca}
resCA_hitdf <- subset(resCA,
   padj <= 0.2 &
   abs(log2FoldChange) > log2(1))
resCA_hits <- setNames(
   sign(resCA_hitdf$log2FoldChange),
   rownames(resCA_hitdf))

resCA_hits
```

The two [signed sets](#signed-set) are combined into a `list` then
visualized with `venndir()` in Figure \@ref(fig:deseq2-example4).

(ref:deseq2-example4) Directional Venn diagram comparing statistical hits from two contrasts analyzed using DESeq2.

```{r deseq2-example4, fig.height=5, fig.width=7, out.width="80%", fig.cap="(ref:deseq2-example4)", fig.alt="(ref:deseq2-example4)"}
setlist <- list(
   "B-A"=resBA_hits,
   "C-A"=resCA_hits)
v <- venndir(setlist,
   legend_headers=c(Set="Contrast", Size="Differential Genes"))
```

To view the genes in the Venn diagram, skip ahead to [Item labels] and
[Signed items], or to extract the genes use `overlaplist(v)`.

