# Venndir Basics

```{r setup-03, include=FALSE}
knitr::opts_chunk$set(
   # fig.retina = 2, out.width = '100%'
   cache=TRUE,
   fig.align="center",
   fig.width=8,
   fig.height=7
)
```

```{r html-latex-color, echo=FALSE}
# custom function to colorize latex output
html_to_latex_color <- function
(tvt,
 verbose=FALSE,
 ...)
{
   #
   tvt2 <- gsub("&nbsp;", " ", gsub("<span style..font-family[^>]+>|</span>", "", tvt))
   # cat(gsub("<", "\n<", tvt2))
   tvl <- unlist(strsplit(gsub("<", "\n<", tvt2), "\n"))
   tvdf <- data.frame(jamba::rbindList(strsplit(gsub(">", ">!", tvl), "!")));
   
   tvl <- lapply(seq_len(nrow(tvdf)), function(irow){
      # compress spaces?
      # tvdf[irow, 2] <- gsub("  ", " ", tvdf[irow, 2]);
      
      tvdf[irow, 2] <- gsub("\u2191", "\\\\uparrow", tvdf[irow, 2])
      tvdf[irow, 2] <- gsub("\u2193", "\\\\downarrow", tvdf[irow, 2])
      tvdf[irow, 2] <- gsub("((\\\\(up|down)arrow)+)", "$\\1$", tvdf[irow, 2])
      if (all(grepl("^<br[^a-z]", tvdf[irow, 1]))) {
         tvdf[irow, 2] <- "\\newline\n";
      } else if (all(grepl("^[ ]*$", tvdf[irow, 2]))) {
         k <- paste0("\\colorbox{white!0}{",
            "\\textcolor{white!0}{",
            tvdf[irow, 2], "}}")
      } else {
         i <- gsub("<span style=.|.>", "", tvdf[irow, 1])
         fg <- gsub("(^|;)color:#([A-F0-9]{6}).*$", "\\2", i)
         # escape the ampersand
         tvdf[irow, 2] <- gsub("([&_])", "\\\\\\1", tvdf[irow, 2]);
         if (grepl("background-color", i)) {
            # \colorbox{yellow}{This text has a yellow background.}
            bg <- gsub("^.*background-color:#([A-F0-9]{6}).*$", "\\1", i)
            k <- paste0("\\colorbox[HTML]{", bg, "}{",
               "\\textcolor[HTML]{", fg,
               "}{", tvdf[irow, 2], "}}")
         } else {
            k <- paste0("\\colorbox{white!0}{",
               "\\textcolor[HTML]{", fg,
               "}{", tvdf[irow, 2], "}}")
            # k <- paste0(
            #    "\\textcolor[HTML]{", fg,
            #    "}{", tvdf[irow, 2], "}")
         }
      }
   })
   # print(tvl)
   tv_text <- jamba::cPaste(sep="", unlist(tvl))
   tv_lines <- jamba::cPaste(sep="", 
      lapply(strsplit(tv_text, "\\\\newline\n"), function(i){
         paste0("\\mbox{\\texttt{", i, "}}\\newline\n")
      }))
   paste0(
      "\\setlength{\\fboxsep}{1pt}\n",
      gsub(" ", "\\\\phantom{M}", tv_lines),
      "\\setlength{\\fboxsep}{8.5pt}\n")
}
```


A Venn diagram seems relatively simple and straightforward.
It uses circles (or sometimes ellipses) to represent the
shared and distinct elements of each set.

Proportional [Euler diagrams](#euler-diagram) use areas proportional to
the number of elements of each set, and are somewhat unpredictable.
<!-- The R package **euler** [@R-eulerr] is used to provide the -->
<!-- best possible Euler diagram, but sometimes it is not possible -->
<!-- to represent all the areas perfectly. -->

An important feature of any good Venn diagram is clear labeling.
Venndir is as much about labeling as it is about the diagram.
Adding *signed labels* only heightens the challenge.

(ref:venndir-manual-label) Three-way proportional Euler diagram, showing the difficulty of labeling each region. Here, the overlap of set_A and set_C has 2 items. The label is placed outside, connected by a line segment.

(ref:venndir-manual-label-alt) Three-way Euler diagram showing the difficulty with placing labels in each overlap region. One region is labeled outside the diagram, with a connecting line segment.

```{r venndir-manual-label, echo=FALSE, out.width="65%", fig.cap="(ref:venndir-manual-label)", fig.alt="(ref:venndir-manual-label-alt)"}
library(venndir)
setlist <- make_venn_test(n_sets=3,
   do_signed=TRUE, concordance=0.7,
   n_items=110,
   sizes=c(20, 30, 10))
v2 <- venndir(setlist, proportional=TRUE, show_labels="ncs",
   do_plot=FALSE,
   setlist_labels=paste("Set", LETTERS[1:3]),
   font_cex=c(1, 1, 0.7) * 1.6,
   circle_nudge=list(set_A=c(0.3, 0.3)), inside_percent_threshold=1)
v2h <- highlight_venndir_overlap(v2, overlap_set="set_A&set_C",
   outerborder="red", outerborder.lwd=2)
plot(v2h,
   expand_fraction=c(1, 1, 1, 1) * -0.1,
   draw_legend=FALSE)

```

Venndir employs a variety of strategies to automate labeling,
and adds approaches to allow for manual adjustments.
Figure \@ref(fig:venndir-manual-label) demonstrates that sometimes
there is no other option but to place a label outside.

Most of this chapter describes basic features of Venndir.
Although some techniques may feel advanced,
I promise it becomes more "advanced" in the next chapter!


## Default Venndir

The following steps begin with a `setlist`, which contains varying
examples for the purpose of demonstrating various features.
To create a `setlist`, refer to [Data Import] for detailed
examples.

### Data for Testing

Most examples use `make_venn_test()` to create a `setlist`.
This function creates simple sets by default,
or signed sets with argument `do_signed=TRUE`.

The arguments help customize useful characteristics:

* `n_items=200`: Overall number of items in the "universe".
* `n_sets=3`: Number of sets.
* `do_signed=FALSE`: Whether to create signed sets.
* `concordance=0.5`: The directional concordance on a scale of -1 to 1.
* `sizes`: Specific sizes for each set, optional.
* `min_size`,`max_size`: Minimum and maximum set size, when `sizes` is NULL.
* `items`: The full universe of items to use, optional.
* `set_names`: Specific set names, optional.

```{r make-setlist}
setlist <- make_venn_test()
```

### Create the Venn 

Given a `setlist`, a Venn diagram is straightforward with `venndir()`.

(ref:default-venn) Default Venn diagram with three sets.

```{r default-venn, out.width="60%", fig.width=5, fig.height=7, fig.cap="(ref:default-venn)", fig.alt="(ref:default-venn)"}
venndir(setlist)
```

The output of `venndir()` is an object with class `Venndir`,
returned invisibly. Returning the object invisibly means that the
object is not printed to screen, but can be assigned to a variable
if desired.

The `Venndir` object contains all the supporting data
required to make a figure, in addition to other relevant data that can
be exported for review, or edited to customize the figure.

A `Venndir` object can be printed which shows a brief summary.

```{r show-venndir}
v <- venndir(setlist, do_plot=FALSE)
v
```

There are many customization options, which will be explored in detail
throughout this user guide.


### Create the Euler

To make a proportional [Euler diagram](#euler-diagram), use argument
`proportional=TRUE`, Figure \@ref(fig:default-euler).

(ref:default-euler) Default proportional Euler diagram with three sets.

```{r default-euler, out.width="70%", fig.width=6, fig.height=7, fig.cap="(ref:default-euler)", fig.alt="(ref:default-euler)"}
venndir(setlist, proportional=TRUE)
```

#### Euler with ellipse

Venndir uses R package **eulerr** [@R-eulerr] to define the
model Euler diagram, which uses circles by default.
However, it can be coerced to use ellipse shapes with
argument `shape="ellipse"`.

Figure \@ref(fig:default-euler-2) compares circular and elliptical shapes,
and illustrates that neither option is perfect. The circular option
shows one empty set

(ref:default-euler-2) Proportional Euler diagram using circle (left) and ellipse (right).

```{r default-euler-2, out.width="50%", fig.ncol=2, fig.width=5, fig.height=7, fig.cap="(ref:default-euler-2)", fig.alt="(ref:default-euler-2)", fig.subcap=c("Circular sets.", "Elliptical sets.")}
setlist <- make_venn_test(sizes=c(33, 18, 69))
v1 <- venndir(setlist,
   vector_method="label",
   proportional=TRUE)
v2 <- venndir(setlist,
   proportional=TRUE,
   vector_method="label",
   shape="ellipse")
```

Euler diagrams are imperfect with three or more sets, which means
there is no guaranteed geometry (with circles or ellipses) that
guarantees the overlap regions will be exactly proportional to
the number of items in each region.

Now is a good time to mention that Euler diagrams are useful but imperfect.

:::: {.tipbox data-latex=""}

**Key Points:**

1. Euler diagrams are only *approximately proportional* for three or more sets.
2. Some overlap regions may be shown which have no items.
3. Some overlap regions may not be shown which contain items.

A diagram with any [hidden overlaps](#hidden-overlaps-glossary) will display a small
footnote symbol $`r "\u2020"`$ in the bottom-right corner.

::::

Points **1** and **2** test one's coping skills:
"When is it *okay* not to be perfect?"

Point **3** is more concerning, and is discussed in detail
in the section [Hidden Overlaps]. Briefly:

* Hidden overlaps are indicated by a footnote.
* Hidden overlaps and are stored in the `Venndir` object.
* Information about "hidden overlaps" can be reviewed 
using: `print(v)`, `footnotes(v)`, `warnings(v)`, or
`overlaplist(v)`.

The footnote is rendered on the bottom-left corner by default,
and can be controlled using arguments in `render_venndir_footnotes()`.
Notably, argument `footnote_style` can be:

* **`'symbol'`**: (default) only the footnote symbol
* **`'footnote'`**: one line for each footnote
* **`'header'`**: Simple summary of footnote marks.

The intention is for the footnote to be subtle, but visible, so
there is clear indication when something should be reviewed in detail.

(ref:default-euler-3) Euler diagrams showing a footnote symbol (left panel), and footnote comment (right panel).

```{r default-euler-3, out.width="50%", fig.ncol=2, fig.width=5, fig.height=7, fig.cap="(ref:default-euler-3)", fig.alt="(ref:default-euler-3)", fig.subcap=c("Footnote symbol.", "Footnote with comment.")}
setlist3 <- make_venn_test(500, n_sets=4)
v3 <- venndir(setlist3,
   draw_legend=FALSE,
   proportional=TRUE)
v3b <- venndir(setlist3,
   footnote_style="footnote",
   footnote_fontsize=12,
   draw_legend=FALSE,
   proportional=TRUE)
```


## Venn Set Colors

The argument `set_colors` is used to define colors for each set.
When it is `NULL`, as by default, it assigns rainbow categorical
colors by calling the **colorjam** [@R-colorjam] R package.

Figure \@ref(fig:default-venn-set-colors) demonstrates user-defined colors (left), and monochrome transparent colors (right).

(ref:default-venn-set-colors) Venn diagram showing custom set colors: red, orange, blue.

```{r default-venn-set-colors, out.width="50%", fig.ncol=2, fig.cap="(ref:default-venn-set-colors)", fig.alt="(ref:default-venn-set-colors)", fig.subcap=c("Custom colors.", "Transparent blue colors.")}
venndir(setlist,
   set_colors=c("firebrick", "orange", "royalblue"))
# blue transparent colors
venndir(setlist,
   poly_alpha=0.4,
   set_colors=c("royalblue", "royalblue", "royalblue"))
```

The background fill color is also adjusted by `poly_alpha` to
control the alpha transparency.
Values range from 0 (fully transparent) to 1 (fully opaque).
The default `poly_alpha=0.6` is mostly opaque.

Figure \@ref(fig:default-venn-set-colors-dark) shows the effect
of using low transparency, which allows some background colors to become dark.
Notice some text labels become white to maintain visual contrast.

(ref:default-venn-set-colors-dark) Venn diagram using opaque colors, which also causes some text labels to become white for visual contrast.

```{r default-venn-set-colors-dark, fig.cap="(ref:default-venn-set-colors-dark)", fig.alt="(ref:default-venn-set-colors-dark)"}
venndir(setlist,
   set_colors=c("firebrick", "orange", "royalblue"),
   poly_alpha=0.9)
```

:::: {.tipbox data-latex=""}

**Note:**

Some R graphics devices do not support alpha transparency,
which can be reviewed with `dev.capabilities()`. It should be run
when the specific graphics device is open, for example after using
`cairo_pdf()` to open a PDF file for output.

::::

Colors can also be customized in any section of the figure,
using techniques described in [Modify Venn Overlaps].

### Consistent Set Colors

When the argument `setlist` is passed to `venndir()`, and no custom
colors are defined with argument `set_colors`, categorical colors are assigned
to each set.

In a situation where `setlist` contains five sets, it is recommended to
provide all five sets even for a figure that uses only a few sets.
The following steps make sure colors are used consistently for each set name.

1. Supply `venndir()` with the complete `setlist`.
2. Use argument `sets` to indicate which entries in `setlist` should be used.
3. Optionally define `set_colors`, `setlist_labels`, and `legend_labels`
for all sets in `setlist`.

Figure \@ref(fig:consistent-colors-1) represents five sets
with specific categorical colors assigned to each set. When
making Venn diagrams with two sets, they should use consistent
colors.

(ref:consistent-colors-1) Five sets are depicted as circles, each with a specific categorical color.

```{r consistent-colors-1, echo=FALSE, fig.align="center", out.width="50%", fig.height=4, fig.width=5.8, fig.cap="(ref:consistent-colors-1)", fig.alt="(ref:consistent-colors-1)"}
setlist1 <- counts2setlist(c(A=5, B=4, C=6, D=3, E=5))
vs1 <- venndir(setlist1, proportional=TRUE, do_plot=FALSE)

jpbox <- bbox_JamPolygon(vs1@jps)
plot(vs1@jps,
   xlim=expand_range(jpbox[1,],
      expand_fraction=-0.3))
```

Figure \@ref(fig:consistent-colors-2) shows the effects in a Venn diagram
with sets A and E, using the respective colors.

(ref:consistent-colors-2) Venn diagram comparing sets A and E. The colors are taken from the five-color palette, using yellow for A, and blue for E.

```{r consistent-colors-2, fig.height=4, fig.width=5, out.width="70%", fig.cap="(ref:consistent-colors-2)", fig.alt="(ref:consistent-colors-2)"}
setlist <- make_venn_test(n_sets=5, n_items=70, set_names=LETTERS[1:5])
v2 <- venndir(setlist,
   sets=c(1, 5))
```

Figure \@ref(fig:consistent-colors-5) shows the desired outcome,
several Venn diagrams using colors consistent for each set.

(ref:consistent-colors-5) Four Venn diagrams using consistent colors for each set.

```{r consistent-colors-5, fig.height=6, fig.width=8, echo=FALSE, out.width="80%", fig.cap="(ref:consistent-colors-5)", fig.alt="(ref:consistent-colors-5)"}
# correct usage
v2ce1 <- venndir(setlist, sets=c(1, 2), do_plot=FALSE)
v2ce2 <- venndir(setlist, sets=c(2, 4), do_plot=FALSE)
v2ce3 <- venndir(setlist, sets=c(1, 3), do_plot=FALSE)
v2ce4 <- venndir(setlist, sets=c(3, 4), do_plot=FALSE)

v2ce1_grobs <- plot(v2ce1, x_inset=grid::unit(-1, "lines"), do_draw=FALSE)
v2ce2_grobs <- plot(v2ce2, x_inset=grid::unit(-1, "lines"), do_draw=FALSE)
v2ce3_grobs <- plot(v2ce3, x_inset=grid::unit(-1, "lines"), do_draw=FALSE)
v2ce4_grobs <- plot(v2ce4, x_inset=grid::unit(-1, "lines"), do_draw=FALSE)

v2ce1_gtree <- attr(v2ce1_grobs, "gtree")
v2ce2_gtree <- attr(v2ce2_grobs, "gtree")
v2ce3_gtree <- attr(v2ce3_grobs, "gtree")
v2ce4_gtree <- attr(v2ce4_grobs, "gtree")

pw2 <- (wrap_elements(v2ce1_gtree) +
   wrap_elements(v2ce2_gtree)) /
   (wrap_elements(v2ce3_gtree) +
   wrap_elements(v2ce4_gtree)) +
   plot_layout(widths=c(1, 1));
pw2
```

Figure \@ref(fig:consistent-colors-4) shows the problem to avoid,
several Venn diagram with colors not consistent for each set.

(ref:consistent-colors-4) Four Venn diagrams using the default colors, illustrating the problem to avoid: Colors are not consistent for each set.

```{r consistent-colors-4, fig.height=6, fig.width=8, echo=FALSE, out.width="80%", fig.cap="(ref:consistent-colors-4)", fig.alt="(ref:consistent-colors-4)"}
# assemble panels with patchwork

v2be1 <- venndir(setlist[c(1, 2)], do_plot=FALSE)
v2be2 <- venndir(setlist[c(2, 4)], do_plot=FALSE)
v2be3 <- venndir(setlist[c(1, 3)], do_plot=FALSE)
v2be4 <- venndir(setlist[c(3, 4)], do_plot=FALSE)

v2be1_grobs <- plot(v2be1, x_inset=grid::unit(-1, "lines"), do_draw=FALSE)
v2be2_grobs <- plot(v2be2, x_inset=grid::unit(-1, "lines"), do_draw=FALSE)
v2be3_grobs <- plot(v2be3, x_inset=grid::unit(-1, "lines"), do_draw=FALSE)
v2be4_grobs <- plot(v2be4, x_inset=grid::unit(-1, "lines"), do_draw=FALSE)

suppressWarnings({
   v2be1_gtree <- attr(v2be1_grobs, "gtree")
   v2be2_gtree <- attr(v2be2_grobs, "gtree")
   v2be3_gtree <- attr(v2be3_grobs, "gtree")
   v2be4_gtree <- attr(v2be4_grobs, "gtree")
})
pw <- (wrap_elements(v2be1_gtree) +
   wrap_elements(v2be2_gtree)) /
   (wrap_elements(v2be3_gtree) +
   wrap_elements(v2be4_gtree)) +
   plot_layout(widths=c(1, 1));
pw
```

Other alternatives to avoid mis-using color:

* Supply argument `set_colors` to `venndir()` with specific colors per set.
* Create Venn with no color, and no background fill.


### Venndir Without Color

It is possible to create a Venndir without color.
For example, some publications require black-and-white output.

Figure \@ref(fig:bw-venn-1) demonstrates Venndir output with no color.

(ref:bw-venn-1) Venndir figure with no color.

```{r bw-venn-1, fig.cap="(ref:bw-venn-1)", fig.alt="(ref:bw-venn-1)"}
v <- venndir(make_venn_test(),
   show_segments=FALSE,
   poly_alpha=0,
   border="black",
   border.lwd=2,
   legend_color_style=c("noborder", "nofill"))
```

:::: {.tipbox data-latex=""}

**Tip:**

Using `poly_alpha=0` also makes signed labels not colored.  
To maintain signed colors, use `poly_alpha=0.01`.  
For more control over signed label colors, see [Custom Sign Label Colors].

::::


## Venndir Labels

### Label Content

The term 'label' may refer to many elements of a Venndir figure.
Of course, the sets have labels, and Venn overlap counts have
labels. When using signed data, there are also signed count labels.
There are also optional labels such as percent overlap, and
even individual item labels. In future, there may be additional
labels such as Jaccard overlap, Kruskal's directional concordance,
or z-score of directionality.

There are so many types of labels, it became unwieldy to assign
a separate function argument for each label, which is why
labels are controlled by one argument: `show_labels`.

Table \@ref(tab:label-table) describes each type of label,
and the letter assigned to each.

(fig:label-table) Summary of recognized options for the argument show labels.

```{r label-table, echo=FALSE}
label_df <- data.frame(check.names=FALSE,
   Label=c("Name",
      "Count",
      "Signed",
      "Percent",
      "Item"),
   Letter=c("'N' or 'n'",
      "'C' or 'c'",
      "'S' or 's'",
      "'P' or 'p'",
      "'i'"),
   Notes=c("The set name provided in 'setlist'.",
      "The count of the number of items in each Venn region. This label is sometimes referred to as 'main count' or 'overlap count'.",
      "The count of the number of items by each observed combination of signs.",
      "The percent of items in each Venn region, related to the total items represented in the diagram.",
      "The item labels represented in each Venn region. Items can only be displayed inside the Venn diagram."
   ))

# prevent kableExtra from loading 'tabu' which is deprecated
options("kableExtra.latex.load_packages"=FALSE)

label_kdf <- knitr::kable(label_df,
   booktabs=TRUE,
   caption="Summary of recognized options for the argument show labels.",
   row.names=FALSE)
label_kdf <- label_kdf |>
   kableExtra::column_spec(1:2, width=c("4em"),
      extra_css="white-space: nowrap;") |>
   kableExtra::column_spec(3, width=c("20em"))
label_kdf <- epub_kdf(label_kdf)
label_kdf
```

Including the respective letter will enable that label.
Using UPPERCASE places the label outside, and using
lowercase places the label inside the Venn diagram.

The default `show_labels="Ncs"` places the Name outside, then
counts and [signed counts](#signed-counts) inside.

:::: {.tipbox data-latex=""}

**Tip:**

A good default to include the percentage is `show_labels="Ncsp"`.
The percentage is always located with the [main counts](#main-counts),
unless [main counts](#main-counts) are not shown.

::::

Currently, item labels can only be placed inside.
It is possible to place item labels in a table outside the Venn
diagram, however it involves advanced techniques.

Figure \@ref(fig:label-info) illustrates common examples for `show_label`,
showing how the components of each label are grouped together.

(ref:label-info) Four examples of various Venndir labels, placed inside or outside each figure. The label components are grouped by location, then organized in a defined way.

```{r label-info, fig.height=4, fig.width=4, out.width="50%", echo=FALSE, fig.ncol=2, fig.cap="(ref:label-info)", fig.alt="(ref:label-info)", fig.subcap=c("show\\_label='Ncps'", "show\\_label='NCPs'", "show\\_label='ncps'", "show\\_label='NCPSi'")}
setlist1 <- make_venn_test(n_sets=2,
   n_items=40,
   set_names=c("Set Label", "B"),
   do_signed=TRUE)

use_fontcex <- 1.2;
use_expand <- c(0.0,0.0,0.0,0.5);
use_xyratio <- 1.8;
label_style <- "lite box"

v <- venndir(setlist1, sets=1,
   do_plot=FALSE,
   show_labels="Ncsp",
   set_colors=c("slateblue1", "gold"),
   setlist_labels=c(`Set Label`="Set Name", B="B"),
   label_style=label_style,
   font_cex=c(1, 1, 0.7) * use_fontcex,
   expand_fraction=c(1,1,1,1)*use_expand,
   segment_distance=0.2)
v@label_df$text[1] <- "{.u c}ount\n{.u p}ercent"
v@label_df$text[2] <- paste(#v@label_df$text[2],
   "\u2191 {.u s}igned")
v@label_df$text[3] <- paste(#v@label_df$text[3],
   "\u2193 {.u s}igned")
v@label_df$venn_label <- "{.u N}ame"
vNcsp <- v;
# plot(vNcsp)

v <- venndir(setlist1, sets=1,
   do_plot=FALSE,
   show_labels="NCsp",
   set_colors=c("slateblue1", "gold"),
   setlist_labels=c(`Set Label`="Set Name", B="B"),
   label_style=label_style,
   font_cex=c(1, 1, 0.7) * use_fontcex,
   expand_fraction=c(1,1,1,1)*use_expand,
   segment_distance=0.2)
v@label_df$text[1] <- "{.u C}ount\n{.u P}ercent"
v@label_df$text[2] <- paste(#v@label_df$text[2],
   "\u2191 {.u s}igned")
v@label_df$text[3] <- paste(#v@label_df$text[3],
   "\u2193 {.u s}igned")
v@label_df$venn_label <- "{.u N}ame"
vNCPs <- v;
# plot(vNCPs)

v <- venndir(setlist1, sets=1,
   do_plot=FALSE,
   show_labels="ncsp",
   set_colors=c("slateblue1", "gold"),
   setlist_labels=c(`Set Label`="Set Name", B="B"),
   label_style=label_style,
   font_cex=c(1, 1, 0.7) * use_fontcex,
   expand_fraction=c(1,1,1,1)*use_expand,
   segment_distance=0.2)
v@label_df$text[1] <- "{.u c}ount\n{.u p}ercent"
v@label_df$text[2] <- paste(#v@label_df$text[2],
   "\u2191 {.u s}igned")
v@label_df$text[3] <- paste(#v@label_df$text[3],
   "\u2193 {.u s}igned")
v@label_df$venn_label <- "{.u n}ame"
vncps <- v;
# plot(vncps)

v <- venndir(setlist1, sets=1,
   do_plot=FALSE,
   show_labels="NCSPi",
   set_colors=c("slateblue1", "gold"),
   setlist_labels=c(`Set Label`="Set Name", B="B"),
   label_style=label_style,
   item_buffer=-0.01, item_cex=0.8, show_items="item", xyratio=2.5,
   font_cex=c(1, 1, 0.7) * use_fontcex,
   expand_fraction=c(1,1,1,1)*use_expand,
   segment_distance=0.2)
v@label_df$text[1] <- "{.u C}ount\n{.u P}ercent"
v@label_df$text[2] <- paste(#v@label_df$text[2],
   "\u2191 {.u S}igned")
v@label_df$text[3] <- paste(#v@label_df$text[3],
   "\u2193 {.u S}igned")
v@label_df$venn_label <- "{.u N}ame"
vNCPSi <- v;
# plot(vNCPSi, xyratio=1.8)
rm(v)

plot(vNcsp,
   expand_fraction=c(1,1,1,1)*use_expand,
   draw_legend=FALSE)
plot(vNCPs,
   expand_fraction=c(1,1,1,1)*use_expand,
   draw_legend=FALSE)
plot(vncps,
   expand_fraction=c(1,1,1,1)*use_expand,
   draw_legend=FALSE)
plot(vNCPSi,
   expand_fraction=c(1,1,1,1)*use_expand,
   xyratio=use_xyratio,
   draw_legend=FALSE)

```


### Overlap Type

The argument `overlap_type` provides different approaches to
summarize directional overlaps. This option determines which
count labels will be displayed in the Venndir figure.

Table \@ref(tab:overlap-type-table) summarizes the options for `overlap_type`.

(fig:overlap-type-table) Summary of recognized overlap types, with a description of the associated output in a Venndir figure.

```{r overlap-type-table, echo=FALSE}
oll <- list(
   "'overlap'"=paste("Only the summary overlap counts are displayed."),
   "'concordance'"=paste("The summary overlap counts are displayed.",
      "Signed counts are tabulated for overlaps involving one direction.",
      "All other signed counts are summarized with 'X' for discordance."),
   "'each'"=paste("The summary overlap counts are displayed.",
      "Signed counts are tabulated for each combination of signs observed."),
   "'agreement'"=paste("The summary overlap counts are displayed.",
      "Signed counts are tabulated based upon agreement or disagreement",
      "of the directional sign.")
)
oldf <- data.frame("overlap_type"=names(oll),
   "Description"=unlist(oll))
kdf <- knitr::kable(row.names=FALSE,
   booktabs=TRUE,
   caption="Summary of recognized overlap types, with a description of the associated output in a Venndir figure.",
   oldf)
kdf <- kdf |>
   kableExtra::column_spec(2, width="25em")
kdf <- epub_kdf(kdf)
kdf
```

* When the input `setlist` is not signed,
`overlap_type='overlap'` is the default.
* When the input `setlist` is signed,
`overlap_type='concordance'` is the default.

Figure \@ref(fig:overlap-type-1) illustrates the four values for
`overlap_type`.

(ref:overlap-type-1) Venn diagrams showing four overlap types recognized: 'overlap' (default for non-signed data), 'concordance' (default for signed data), 'each', and 'agreement'.

(ref:overlap-type-1a) overlap_type='overlap'

(ref:overlap-type-1b) overlap_type='concordance'

(ref:overlap-type-1c) overlap_type='each'

(ref:overlap-type-1d) overlap_type='agreement'

```{r overlap-type-1, echo=FALSE, out.width="50%", fig.height=6, fig.width=6, fig.cap="(ref:overlap-type-1)", fig.alt="(ref:overlap-type-1)", fig.ncol=2, fig.subcap=c("(ref:overlap-type-1a)", "(ref:overlap-type-1b)", "(ref:overlap-type-1c)", "(ref:overlap-type-1d)")}
setlist <- make_venn_test(do_signed=TRUE)

v1 <- venndir(setlist, overlap_type='overlap')
v2 <- venndir(setlist, overlap_type='concordance',
   x_inset=grid::unit(-1, "lines"))
v3 <- venndir(setlist, overlap_type='each',
   x_inset=grid::unit(-1, "lines"))
v4 <- venndir(setlist, overlap_type='agreement',
   x_inset=grid::unit(-1, "lines"))

```

### Signed Label Placement

Signed count labels are placed beside main count labels by default,
however they can be placed below main count labels.

The argument `template` controls the signed label placement,
and there are two options, illustrated in Figure \@ref(fig:template-wide).

1. `template='wide'` (default) labels [signed counts](#signed-counts) beside
main overlap counts.
2. `template='tall'` labels [signed counts](#signed-counts) below
main overlap counts.

(ref:template-wide) Venn diagram with signed counts labeled beside overlap counts (left) using `template='wide'`; and below overlap counts (right) using `template='tall'`.

```{r template-wide, out.width="50%", fig.ncol=2, fig.height=6, fig.width=6, fig.cap="(ref:template-wide)", fig.alt="(ref:template-wide)", fig.subcap=c("template='wide'", "template='tall'")}
setlist3 <- make_venn_test(n_sets=3, do_signed=TRUE)
vt1 <- venndir(setlist3)
vt2 <- venndir(setlist3, template="tall")
```



### Visual Styles

By default, Venn count and set labels are drawn without any particular
color shading or border. The argument `label_style` is used to enable
background color fill, and optional border.

The recognized keywords for `label_style`:

* `"lite"` - light background
* `"shaded"` - semi-transparent shaded color background
* `"fill"` - full color background
* `"box"` - draw a box as a border around the label

A straightforward example is shown in Figure \@ref(fig:label-style-0)
using `label_style="lite box"` with [signed counts](#signed-counts).

(ref:label-style-0) Venn diagram showing signed counts, using label style 'lite box'.

```{r label-style-0, fig.height=8, fig.width=6, fig.alt="(ref:label-style-0)", fig.cap="(ref:label-style-0)"}
setlistS <- make_venn_test(do_signed=TRUE)
vS <- venndir(setlistS, label_style="lite box")
```

Figure \@ref(fig:label-style-1) illustrates several possible styles.
In general, when using a fill color with any of the options
'lite', 'shaded', 'fill', it may be preferred to add 'box' to include
a visual border. For example, try 'lite box' or 'shaded box'.

(ref:label-style-1) Venn diagrams showing the different label style options in each of four panels.

(ref:label-style-1a) label_style='lite'

(ref:label-style-1b) label_style='shaded'

(ref:label-style-1c) label_style='fill'

(ref:label-style-1d) label_style='lite box'

```{r label-style-1, fig.ncol=2, out.width="50%", fig.height=4, fig.width=5, echo=FALSE, fig.alt="(ref:label-style-1)", fig.cap="(ref:label-style-1)", fig.subcap=c("(ref:label-style-1a)", "(ref:label-style-1b)", "(ref:label-style-1c)", "(ref:label-style-1d)")}
setlist <- make_venn_test(n_sets=3)
v1 <- venndir(setlist, sets=c(1, 3), label_style="lite")
v2 <- venndir(setlist, sets=c(1, 3), label_style="shaded")
v3 <- venndir(setlist, sets=c(1, 3), label_style="fill")
v4 <- venndir(setlist, sets=c(1, 3), label_style="lite box")
```

### Nudge a Venn Label

To nudge, that is to reposition, one label in a Venndir diagram,
use `nudge_venndir_label()`.

Note that `venndir()` will define a label position for every Venn overlap,
and defines a position 'inside' and 'outside' the Venn region.

:::: {.tipbox data-latex=""}

**Exception:**

Sometimes a proportional Euler diagram cannot represent
every overlap, due to geometric constraints.
In this case, not every overlap will be assigned a label position.

For a discussion of that issue, see Section \@ref(hidden-overlaps)
[Hidden Overlaps].

::::

To nudge a label, define the label to nudge using these two arguments:

* `set` - the overlapping region to nudge
* `label_location` - the inside or outside label associated with that set

The label adjustment uses two coordinates. Units are proportional
to the overall Venndir plot region, where `1` is the full width
or height of the plot, whichever is larger.

* `x_offset`
* `y_offset`

The process requires an existing `Venndir` object.
In the example below, a simple 3-way Venn diagram is stored
in variable `v`.
The label for `'set_C'` is moved left (decreasing the x coordinate),
and up (increasing the y coordinate).

```{r nudge-label-1a, eval=FALSE}
# default Venn
setlist <- make_venn_test()
v <- venndir(setlist)

# nudge 'set_C' up-and-left
v2 <- nudge_venndir_label(v,
   x_offset=-0.45, y_offset=0.25,
   set="set_C",
   label_location="outside")
plot(v2)
```

Figure \@ref(fig:nudge-label-1) shows the Venn with
default labels (left), and with the label `'set_C'` adjusted (right).

(ref:nudge-label-1) Default 3-way Venn diagram (left), with label `'set_C'` moved up and to the left (right).

(ref:nudge-label-1-1) Default Venn.

(ref:nudge-label-1-2) Adjusted set_C label.

```{r nudge-label-1, echo=FALSE, fig.ncol=2, fig.height=6, fig.width=6, out.width="50%", fig.cap="(ref:nudge-label-1)", fig.alt="(ref:nudge-label-1)", fig.subcap=c("(ref:nudge-label-1-1)", "(ref:nudge-label-1-2)")}
setlist <- make_venn_test()
v <- venndir(setlist, do_plot=FALSE)
plot(v)

v2 <- nudge_venndir_label(v,
   x_offset=-0.45, y_offset=0.25,
   set="set_C",
   label_location="outside")
plot(v2)
```

Notice that after the label is adjusted, the new `Venndir` object `v2`
can be visualized using `plot()`.


### Label Placement Rules

For each Venndir figure, sets are each represented by a circle
or ellipse. It should be "easy" to label each circle, right?
Experience has shown this to be more difficult than expected.

(ref:label-rules) Two examples of with very challenging labels.

(ref:label-rules-1) Fully nested sets.

(ref:label-rules-2) Partially nested sets.

```{r label-rules, echo=FALSE, fig.ncol=2, fig.height=4, fig.width=5, out.width="50%", fig.cap="(ref:label-rules)", fig.alt="(ref:label-rules)", fig.subcap=c("(ref:label-rules-1)", "(ref:label-rules-2)")}
venndir(counts2setlist(c(A=7, "A&B"=3, "A&B&C"=1)),
   proportional=TRUE,
   innerborder="white", innerborder.lwd=0.7, outerborder=NA,
   vector_method="label",
   segment_buffer=-0.3,
   legend_color_style="whiteborder",
   circle_nudge=list(C=c(-0.5, -0.2), B=c(-0.7, -0.1)),
   segment_method="vector")

venndir(counts2setlist(c(A=7, "A&B"=3, "A&B&C"=2, "A&C"=1)),
   proportional=TRUE,
   innerborder="white", innerborder.lwd=0.7, outerborder=NA,
   vector_method="label",
   segment_buffer=-0.3,
   legend_color_style="whiteborder",
   circle_nudge=list(C=c(-0.5, -0.7), B=c(-0.5, -0.7)), rotate_degrees=-40,
   segment_method="vector")

```

In Figure \@ref(fig:label-rules), where should the Set label be placed?
Where should the line segments connect?

**Specific rules for placing labels in the Venn figure:**

1. **Set labels** are directed to the most specific region in the figure.

   * For Venn diagrams, Set labels refer to the region
   "unique to set_A", with no other overlapping sets.
   * For Euler diagrams, Set labels refer to the region "unique to set_A"
   *if it exists*. Otherwise the Set refers to a region
   containing *"set_A"* with the fewest overlapping sets.
   * In Figure \@ref(fig.label-rules) (left) Set **'C'** refers to
   the region inside B and C.
   * In Figure \@ref(fig.label-rules) (right) Sets **'B'** and **'C'**
   each refer to regions as unique to B and C as possible, respectively.

2. **Count labels** will only be associated with the specific
overlap region.

   * If the specific overlap region does not exist in the figure,
   the count label is not shown. Even though it may seem obvious,
   sometimes a count label cannot be shown. See [Hidden Overlaps].
   * If a count label is placed outside, a line segment will
   connect the label to the corresponding region.
   * The argument `inside_percent_threshold` may be used to
   place a label outside when the region is small.

3. **Set and Count labels** may be grouped together when both are inside
or outside.

   * When placed inside, a Set label will be grouped with the count label
   to which the Set label refers, as described in Rule 1.
   * When placed outside, a Set label will only appear with a count label
   that is specific to that Set.

4. **Percentage labels** are shown together with [main counts](#main-counts)
when [main counts](#main-counts) are shown.

   * Percentages are calculated directly from [main counts](#main-counts),
   and as a result it is most intuitive for them appear together.
   * However, if [main counts](#main-counts) are hidden, the percentage
   may be placed inside or outside.
   
5. **Items labels** are only displayed inside, see [Item Labels].

   * Count labels are "moved outside" if they were enabled inside and items
   are to be displayed.
   * When the number of items exceeds `max_items` then items are hidden,
   and the count labels may be displayed in their place.
   * Set names may be displayed inside, together with item labels,
   however no effort is made to prevent overlapping labels.
   Set labels can be manually nudged, see [Nudge a Venn Label].


## Item Labels

To display items inside a Venn diagram, add `"i"` to the argument
`show_labels`. For example `show_labels="Ni"` will display the set Name
outside, and items inside.

I never thought I'd be able to place item labels in a Venn diagram, much
less expect it, and eventually *need* it.
Item labeling has become one of my favorite features.

Adding item labels also ticks the box, so to speak, of an important
[data visualization](#data-visualization) paradigm:
Answering the very next question. For a Venn diagram:  
"What are those?"

Figure \@ref(fig:items-1) illustrates an example inspired by a
[published figure](https://www.researchgate.net/figure/Venn-diagram-of-genes-associated-with-susceptibility-to-PAD-CAD-and-stroke-identified_fig2_356459049)
[@Salybekov_2021].

(ref:items-1) Venn diagram depicting genes associated with susceptibility to two diseases. The diagram shows the number of genes in each region: 2, 3, and 2, but does not show the genes.

```{r items-1, echo=FALSE, out.width="80%", fig.cap="(ref:items-1)", fig.alt="(ref:items-1)"}
itemlist <- list(
   Stroke=c("COL4A12", "CHRNA3", "PITX2", "HLA-B", "ABO"),
   PAD=c("HLA-B", "IL-6", "F5", "CHRNA3", "ABO")
)
venndir(itemlist,
   font_cex=1.3,
   set_colors=c("#FFC700", "#12E0E6"), poly_alpha=0.8,
   setlist_labels=c("{#33647A STROKE}", "{#33647A PAD}"),
   alias=c(Stroke="Stroke", PAD="PAD"),
   legend_labels=c("Stroke Susceptibility", "Peripheral Arterial Disease"),
   fontfamily="Proxima Nova",
   fontfaces=list(overlap="plain", count="plain", signed="plain"),
   outerborder.lwd=5, innerborder.lwd=0,
   outerborder="#F5F5F5BB")
```

*(What is the very next question?)*  
**"Which genes are in each region?"**

(ref:items-2) The same Venn diagram now showing genes as item labels inside the diagram, answering *the very next question*.

```{r items-2, echo=FALSE, out.width="80%", fig.cap="(ref:items-2)", fig.alt="(ref:items-2)"}
venndir(itemlist,
   font_cex=1.3,
   show_labels="Ni",
   item_cex=c(2, 2, 1.8), xyratio=1.5,
   set_colors=c("#FFC700", "#12E0E6"), poly_alpha=0.8,
   setlist_labels=c("{#33647A STROKE}", "{#33647A PAD}"),
   fontfamily="Proxima Nova",
   fontfaces=list(overlap="plain", count="plain", signed="plain"),
   outerborder.lwd=5, innerborder.lwd=0,
   outerborder="#F5F5F5BB")
```

Figure \@ref(fig:items-2) represents the first example thus far of
[Figure Boosting], wherein the goal is to re-create published
figures using Venndir.


### Overview of item labeling

There is much utility in adding item labels into a Venn diagram,
but most involve the critical word *'if'*.
For example, "It would be helpful to add labels _if_ they fit."
Or "*if* the labels are legible."

And so there are many options associated with item labels.

Before making changes, it is helpful to understand the process.

1. Items are associated with each Venn overlap region.
   * More than `max_items` items will not be displayed.

1. The type of item label is defined using argument `show_items`.
   * "item": '**ACTB**',
   * "sign": '**$`r "↑"`$**'
   * "sign item": '**$`r "↑ ACTB"`$**'

1. Label coordinates are arrayed across the Venn region.
   * `xyratio` controls the x:y ratio when spacing points
   * `label_method` controls whether to use up/down offset, or row-wise
   column layout

1. The item font size is adjusted per region.
   * The adjustment uses the number of labels, and the area
   of the region relative to the total area.
   
1. Font colors are adjusted for contrast.
   * `make_color_contrast()` determines whether to use light or dark text
   for each background color.

1. Minor "jittering" is applied to each label
   * The effect color, size, and angle of labels are altered
   to help distinguish each label.
   * `jitter_cex=0.04`
   * `jitter_color=0.07`
   * `jitter_degrees=0`

1. Items are sorted by sign, then by name.
1. Items are rendered using **marquee** [@R-marquee].

Note that no special effort is made to prevent overlaps between labels,
nor to prevent labels from extending outside the region.
These features may be implemented in future.


### Venn Memes

`venn_meme()` provides a streamlined approach to displaying items.
The data input process was described in [Data Import],
in the section [Items].

The `venn_meme()` function itself calls `venndir()` with some pre-defined
default settings, otherwise it can be customized the same way as
with `venndir()`.

[Figure Boosting], for many cases, is easily carried out
with `venn_meme()` alone.
provides the most convenient starting point.

A simple example of a 3-way Venn meme is shown in Figure \@ref(fig:meme-bix).

(ref:meme-bix) Venn diagram created using `venn_meme()` showing 'Biology', 'Computer Science', and 'Stats' coming together to represent the field of Bioinformatics.

```{r meme-bix, fig.height=8, fig.width=8, fig.cap="(ref:meme-bix)", fig.alt="(ref:meme-bix)"}
bix <- list(
   "Biology",
   "Computer\nScience",
   "Stats",
   "Computational\nBiology",
   "Data\nScience",
   "Biostatistics",
   "Bioinformatics")
venn_meme(bix,
   fontfamily="Futura")
```

Another fun example involves the perils of social media,
and being a sports fan. This example also assigns custom
`"white"` colors for `innerborder` and `outerborder` to
give it a clean look.

A white border can be a nice change from the default dark border,
since it does not conflict with dark text.

(ref:meme-doomscroll) Venn meme created with elements of happiness and rage, doom-scrolling social media, and being a sports fan. It also demonstrates use of a thin white border.

(ref:meme-doomscroll-alt) Venn meme created with elements of happiness and rage, doom-scrolling social media, and being a sports fan.

```{r meme-doomscroll, fig.height=8, fig.width=8, fig.cap="(ref:meme-doomscroll)", fig.alt="(ref:meme-doomscroll-alt)"}
doom <- list(
   "short-lived\nhappiness",
   "prolonged\nsuffering",
   "sudden\nrage",
   "eating\ntoo much\nspicy\nfood",
   "stubbing\nyour toe,\ntwice",
   "doom-\nscrolling\nsocial\nmedia",
   "being\na sports\nfan"
)
venn_meme(doom,
   outerborder="white", innerborder="white",
   fontfamily="Optima", item_cex_factor=0.9,
   set_colors=c("gold", "royalblue", "firebrick3"))
```

### Item Label Basics

By default, labels are arranged, sized relative to each region, and displayed.
Items are displayed by adding `"i"` to the argument `show_labels`, for
example `show_labels="Ni"`. See [Label Content] for a more
thorough description.

(ref:item-basics-1) Venn diagram showing item labels using default settings.

```{r item-basics-1, echo=FALSE, out.width="80%", fig.cap="(ref:item-basics-1)", fig.alt="(ref:item-basics-1)"}
setlist <- make_venn_test(n_items=41)
venndir(setlist,
   item_cex_factor=0.8,
   show_labels="Ni")
```

#### Items inside, counts outside

When items are displayed inside, counts cannot also be displayed inside.
However, counts may be displayed outside by adding UPPERCASE `"C"` to
argument `show_labels`.

(ref:item-basics-2) Venn diagram showing item labels inside, and count labels outside.

```{r item-basics-2, echo=FALSE, out.width="80%", fig.cap="(ref:item-basics-2)", fig.alt="(ref:item-basics-2)"}
setlist <- make_venn_test(n_items=41)
venndir(setlist,
   item_cex_factor=0.8,
   expand_fraction=c(0.1, 0, 0.1, 0),
   show_labels="NCi")
```

#### Signed Items

When the input `setlist` contains signed data, the items by default
will display the sign and item together. This behavior is defined
by the argument `show_items="sign item"`.

Conversely, item labels will not include the sign, if either:
* the input `setlist` does not contain signed data, or
* `overlap_type="overlap"` is defined, which ignores signed data.

(ref:venn-signed-1) Venn diagram with signed data, showing item labels inside each region. The items now have directional arrows beside them, and the item labels utilize color to indicate the directionality.

```{r venn-signed-1, fig.cap="(ref:venn-signed-1)", fig.alt="(ref:venn-signed-1)"}
setlist_signed <- make_venn_test(n_items=41, do_signed=TRUE)
venndir(setlist_signed,
   item_cex_factor=0.7,
   x_inset=grid::unit(-1, "lines"),
   show_labels="Ni")
```

Note that signed item labels are colorized consistent with the
sign, as defined by argument `curate_df`. For a detailed
discussion of signs, and how to customize the sign and corresponding
color, see [Custom Signs].


#### Signs only

When there is a large number of items, one alternative is to display
the sign. This option is effective partly because the color is also used,
and has the effect of "filling the space" with proportional amount of color.

To display only the sign for each item, use `show_items="sign"`.

When showing only the sign, it may be useful to increase the font size.
The argument `item_cex_factor` adjusts item fonts in each region, and
is multiplied by the automated sizing based on the number of items and
area in each region.

(ref:venn-sign-only) Venn diagram showing only the signs of the items filling each overlap region.

```{r venn-sign-only, fig.cap="(ref:venn-sign-only)", fig.alt="(ref:venn-sign-only)"}
setlist_signed <- make_venn_test(n_items=850, do_signed=TRUE)
venndir(setlist_signed,
   show_labels="Ni",
   show_items="sign",
   x_inset=grid::unit(-5, "lines"),
   item_cex_factor=1.2)
```


### Item layout adjustments

There are three main points for adjusting the layout of item labels.

1. `xyratio`: Defines the width-to-height ratio, default 1.1.
   * Useful to increase xyratio for wider labels, or to enforce
   "column" type layout.
2. `label_method`: Determines whether labels alternate up/down per row.
   * `label_method="offset"` (default) uses a slight 'offset' on each row,
   resulting in somewhat triangular label placement. This approach
   is more effective at packing labels into the available space.
   * `label_method="columns"` does not apply an offset, which helps
   align items in neatly organized rows. This option is useful when also
   using a higher value for `xyratio`.
3. `item_buffer`: Controls the buffer zone around the region border.

(ref:venn-layout-1) Four Venn diagrams, showing examples using different values for 'xyratio' and 'label_method'. Higher 'xyratio' values tend to become more column-oriented.

(ref:venn-layout-1a) xyratio=1.1 (default)

(ref:venn-layout-1b) xyratio=1.5

(ref:venn-layout-1c) xyratio=2.0

(ref:venn-layout-1d) xyratio=2.5 in columns

```{r venn-layout-1, fig.ncol=2, out.width="50%", echo=FALSE, fig.height=5, fig.width=7, fig.cap="(ref:venn-layout-1)", fig.alt="(ref:venn-layout-1)", fig.subcap=c("(ref:venn-layout-1a)", "(ref:venn-layout-1b)", "(ref:venn-layout-1c)", "(ref:venn-layout-1d)")}
setlist_signed <- make_venn_test(n_items=150,
   item_prefix="id",
   do_signed=TRUE)
v1 <- venndir(setlist_signed,
   draw_legend=FALSE,
   expand_fraction=-0.15,
   show_labels="i")
v2 <- venndir(setlist_signed,
   draw_legend=FALSE,
   expand_fraction=-0.15,
   xyratio=1.5,
   show_labels="i")
v3 <- venndir(setlist_signed,
   draw_legend=FALSE,
   expand_fraction=-0.15,
   xyratio=2,
   show_labels="i")
v4 <- venndir(setlist_signed,
   draw_legend=FALSE,
   expand_fraction=-0.15,
   xyratio=2.5, label_method="columns",
   show_labels="i")
```


### Item buffer

Two arguments are used during item labeling, both of which are motivated
to help place typical text labels inside the respective Venn overlap
region:

1. The argument `item_buffer` is used to impose a reasonable distance
between item labels and the border of the region to be labeled.
Units are relative to the fraction of area that would shrink the region
to zero.
2. The argument `width_buffer` adds a small buffer to the left and right
sides of each region, which is intended to correct for wide text labels.

The default `item_buffer = -0.15` is negative in order to shrink
the region by this relative amount, and `-0.15` shrinks the
region by 15% relative to the available buffer size for the
region.

Note that `item_buffer = -0.15` (15% buffer) is a slightly different
size for each region.

(ref:item-buffer-1) Four Venn diagrams showing the effects of adjusting 'item_buffer'.

(ref:item-buffer-1a) item_buffer = -0.15 (default)

(ref:item-buffer-1b) item_buffer = -0.3

(ref:item-buffer-1c) item_buffer = 0.0

(ref:item-buffer-1d) item_buffer = 0.1

```{r item-buffer-1, echo=FALSE, fig.ncol=2, out.width="50%", fig.height=5, fig.width=7, fig.cap="(ref:item-buffer-1)", fig.alt="(ref:item-buffer-1)", fig.subcap=c("xyratio=1.1", "xyratio=1.5", "xyratio=2.0", "xyratio=2.5 in columns"), fig.subcap=c("(ref:item-buffer-1a)", "(ref:item-buffer-1b)", "(ref:item-buffer-1c)", "(ref:item-buffer-1d)")}
setlist_signed <- make_venn_test(n_items=1500,
   item_prefix="id",
   do_signed=TRUE)
v1 <- venndir(setlist_signed, show_labels="i",
   expand_fraction=-0.15,
   draw_legend=FALSE,
   show_items="sign")
v2 <- venndir(setlist_signed, show_labels="i",
   expand_fraction=-0.15,
   item_buffer=-0.3,
   draw_legend=FALSE,
   show_items="sign")
v3 <- venndir(setlist_signed, show_labels="i",
   expand_fraction=-0.15,
   item_buffer=0.0,
   draw_legend=FALSE,
   show_items="sign")
v4 <- venndir(setlist_signed, show_labels="i",
   expand_fraction=-0.15,
   item_buffer=0.1,
   draw_legend=FALSE,
   show_items="sign")
```


## Fonts and Font Sizes

If the most challenging aspects of Venndir were the intricacies
of placing labels, surely the next big challenge
was ensuring any chosen font rendered correctly.

Particularly for signed count labels, the default output
uses Unicode up and down arrows: $`r "\u2191"`$ and $`r "\u2193"`$.
These Unicode characters have caused the most heartache,
for various reasons related to Unicode font support and
computer platform locale.

Most* font rendering issues are handled smoothly by the combination of
[marquee](https://marquee.r-lib.org),
[systemfonts](https://systemfonts.r-lib.org), and
[textshaping](https://textshaping.r-lib.org).
These R packages strive to "make it work" by finding each
character glyph in the specified font, or by using a suitable font
substitution.

Particularly for PDF output, it may be necessary to embed
the font into the PDF file using `grDevices::embedFonts()` or
`grDevices::embedGlyphs()`. Using `cairo_pdf()` should embed
fonts automatically, but has rare exceptions.

(ref:marquee-1) Most fonts 'just work' thanks to R package marquee, with enough flexibility to keep the focus on effective and artistic choices, and away from any annoyances.

```{r marquee-1, echo=FALSE, fig.height=2.0, out.width="100%", fig.alt="(ref:marquee-1)", fig.cap="(ref:marquee-1)"}
use_text <- paste(collapse="\n",
   c("### End Result",
      "Most fonts should 'just work' ",
      "with enough flexibility to keep the focus on",
      "effective and artistic choices, not the annoyances. 🎉  ",
      "{.darkorange4 (This text was rendered by marquee, using the 'Optima' font.)}"))
use_style <- marquee::classic_style(
   lineheight=1.1,
   base_size=14,
   body_font="Optima",
   header_font="TrebuchetMS Flex",
   code_font="Andale Mono")
grob <- marquee::marquee_grob(use_text, style=use_style)
grid::grid.newpage()
grid::grid.draw(grob)

```

### Font Sizes

The default font size used in Venndir is `16` points, and is adjusted
with the argument `font_cex`. Default `font_cex=c(1, 1, 0.75)` therefore
defines values `16, 16, 12`, which are applied in order as follows:

* Set label
* Count label
* Signed count label(s)

The text is defined with a fixed font size, and is therefore
independent of the size of the figure. If the figure itself is drawn
on a small graphics device, the font sizes will remain the same,
so some care should be taken to select font sizes appropriate for
the output figure dimensions.

Ideally, the output device dimensions should be defined with fixed units,
such as inches ('in'), or centimeters ('cm'), so that the final
figure is clearly defined.

The argument `font_cex` has specific behavior based upon the number
of values supplied in the argument:

* **`font_cex` with one value**: It is multiplied by `c(1, 1, 0.75)`
to be an easy adjustment to the typical defaults. Thus `font_cex=2` will
double all font sizes.
* **`font_cex` with two values**: The second value is repeated to make a vector
of three values. This vector is multiplied by `c(1, 1, 0.75)`, again as
convenient way to adjust Set and Count labels
proportional to their default values.  
`font_cex=c(2, 1)` will double the Set label, keeping Count and Signed
at their defaults.
* **`font_cex` with three values**: Three values are used directly.

The adjusted `font_cex` is multiplied by `16` to define the font point size.
Several examples are shown in Figure \@ref(fig:font-size-1).

(ref:font-size-1) Venn diagram showing modifications to font sizes by adjusting the argument `font_cex`.

(ref:font-size-1-1) Default

(ref:font-size-1-2) font_cex=1.2

(ref:font-size-1-3) font_cex=c(2, 1)

(ref:font-size-1-4) font_cex=c(2, 1.3, 0.8)

```{r font-size-1, echo=FALSE, out.width="50%", fig.height=5, fig.width=7, fig.ncol=2, fig.cap="(ref:font-size-1)", fig.alt="(ref:font-size-1)", fig.subcap=c("(ref:font-size-1-1)", "(ref:font-size-1-2)", "(ref:font-size-1-3)", "(ref:font-size-1-4)")}
v <- venndir(make_venn_test(do_signed=TRUE),
   expand_fraction=c(0.05, -0.15, -0.05, 0.05),
   x_inset=grid::unit(-5, "lines"))

v <- venndir(make_venn_test(do_signed=TRUE),
   font_cex=1.2,
   expand_fraction=c(0.05, -0.15, -0.05, 0.05),
   x_inset=grid::unit(-5, "lines"))

v <- venndir(make_venn_test(do_signed=TRUE),
   font_cex=c(2, 1),
   expand_fraction=c(0.05, -0.15, -0.05, 0.05),
   x_inset=grid::unit(-5, "lines"))

v <- venndir(make_venn_test(do_signed=TRUE),
   font_cex=c(2, 1.3, 0.8),
   expand_fraction=c(0.05, -0.15, -0.05, 0.05),
   x_inset=grid::unit(-5, "lines"))
```

The last example (bottom-right panel) was created using the following code:

```{r font-size-2, eval=FALSE}
v <- venndir(make_venn_test(do_signed=TRUE),
   font_cex=c(2, 1.3, 0.8))
```

### Overall Font Family

The overall font typeface used by `venndir()` is defined with
the argument `fontfamily`, for which the default is `fontfamily="Arial"`.

See [Available Font Families] to review other fonts available.

Figure \@ref(fig:family-1) uses the R 'serif' font, with `fontfamily='serif'`.
.

(ref:family-1) Venn diagram showing two sets, with all labels shown using a serif font, such as 'Times' or 'Times New Roman'.

```{r family-1, fig.height=6, fig.width=7, fig.cap="(ref:family-1)", fig.alt="(ref:family-1)"}
setlist <- make_venn_test(n_sets=3, do_signed=TRUE)
v <- venndir(setlist,
   main="fontfamily='serif'",
   fontfamily="serif")
```

Note that all the labels use 'serif' now, including in the legend.

Similarly, for Venn diagrams which show item labels,
the item font is adjusted with `fontfamily`.
This type of plot can be created using `venndir()`
with argument `show_labels="Ni"`, to place set name 'N' outside,
and item labels 'i' inside.

(ref:family-2) Venn diagram showing Two sets 'Dogs' and 'Cats' with some attributes unique to each, and shared by both. The item labels use the 'Optima' font.

```{r family-2, fig.height=5, fig.width=7, fig.cap="(ref:family-2)"}
petlist <- list(
   Dog=c("Needs walks", "Barks", "Round eyes"),
   Cat=c("Freely roams", "Meows", "Thin pupils"),
   "Dog&Cat"=c("Furry", "Has claws", "Lovable"))
petsetlist <- overlaplist2setlist(petlist)

petv <- venndir(petsetlist,
   font_cex=1.5,
   item_cex_factor=1.0,
   show_labels="Ni",
   draw_legend=FALSE,
   show_segment=FALSE,
   keep_item_order=TRUE,
   item_buffer=0, xyratio=1.7,
   fontfamily="Optima")

```

The example above includes some other useful customizations, discussed
in detail later in the book. For now, here is a brief description:

* `font_cex=1.5` enlarges the set name font.
* `draw_legend=FALSE` hides the color legend.
* `show_segment=FALSE` hides the line segments.
* `keep_item_order=TRUE` maintains the original order of item labels,
therefore not sorting items alphabetically.
* `item_buffer=0` reduces the buffer zone of each Venn region before
determining item label positions.
* `xyratio=2` increases the x:y ratio during item label placement,
higher values tend to produce single column output.

Item labels can be displayed using `venn_meme()`.

(ref:family-3) Venn diagram showing Two sets 'Dogs' and 'Cats', this time created using `venn_meme()`. The item labels use the 'Optima' font.

```{r family-3, fig.cap="(ref:family-3)", fig.alt="(ref:family-3)"}
petlist <- list(
   Dog=c("Needs walks", "Barks", "Round eyes"),
   Cat=c("Freely roams", "Meows", "Thin pupils"),
   "Dog&Cat"=c("Furry", "Has claws", "Lovable"))
venn_meme(petlist,
   font_cex=2,
   item_cex_factor=1.0,
   item_buffer=0, xyratio=1.7,
   fontfamily="Optima",
   show_labels="Ni")
```


### Individual Font Families

For more control over specific fonts, use argument `fontfamilies`.
This argument expects a `list` with these elements:

* overlap - fontfamily to use for set or overlap labels
* count - fontfamily to use for main count labels
* signed - fontfamily to use for signed count labels

Any value not customized will use the default `fontfamily` value.

This example shows distinct fonts for each element, in fact it shows
two fonts for "count" which corresponds to the [main counts](#main-counts), and
the percentage of counts, in order.

(ref:fontfamilies-1) Venn diagram showing distinct font families used for each type of label: overlap, count, and signed.

```{r fontfamilies-1, fig.height=6.5, fig.width=7, fig.cap="(ref:fontfamilies-1)", fig.alt="(ref:fontfamilies-1)"}
setlist <- make_venn_test(do_signed=TRUE)
venndir(setlist,
   show_labels="Ncps",
   main="Main Title (Times New Roman)",
   fontfamily="Times New Roman",
   y_inset=grid::unit(1, "lines"),
   expand_fraction=c(0.15, 0, 0, 0),
   fontfamilies=list(overlap="Optima",
      count=c("Arial Black", "Times"),
      signed="Arial Narrow"))
```

The arguments to `assemble_venndir_label()` also include `fontfaces` which
may also be customized to control whether each font is 'plain', 'bold',
'italic', or 'bold.italic' for example. These customizations are rather
advanced, and are discussed in more detail later.


### Available Font Families

Before getting into details with fonts, some useful terminology should
be defined. Thomas Pederson wrote an excellent overview of font
terms specific to R in
[Typography and R](https://systemfonts.r-lib.org/articles/fonts_basics.html),
as part of the **systemfonts** [@R-systemfonts] package.

To paraphrase, the commonly used term "font" typically refers to a 'typeface',
for example 'Arial' and 'Helvetica' are each typefaces.
The typeface may have several variations that may include width
(normal, bold, or thin), and style (normal, italic, or oblique).
Each variation represents its own "font" within the typeface.
Some typefaces such as 'Helvetica Neue' may have 18 or more
individual fonts!

With these terms in mind, R refers to a typeface with either the term
`'fontfamily'` or `'family'`, and this convention is followed
in Venndir.

The [systemfonts](https://systemfonts.r-lib.org) R package [@R-systemfonts]
is used by Venndir, by way of the [marquee](https://marquee.r-lib.org)
R package [@R-marquee].

A function `subset_systemfonts()` may be helpful to browse the available
fonts on the system running R. The following example demonstrates how
to summarize available fonts for 'Helvetica Neue', shown in
Table \@ref(tab:systemfonts-1).

```{r systemfonts-1-show, eval=FALSE}
subset_systemfonts(grepl("Helvetica Neue", family))
```

(fig:systemfonts-1) Partial summary of 'Helvetica Neue' fonts via systemfonts.

```{r systemfonts-1, echo=FALSE}
scols <- c("name", "family", "style", "weight", "width", "italic")[1:4]
sdf <- head(
   subset_systemfonts(grepl("Helvetica Neue", family))[, scols, drop=FALSE],
   9)

# Note 'scale_down' actually zooms the table to fit the margins
kdf <- knitr::kable(sdf,
   caption=paste("Partial summary of 'Helvetica Neue' fonts via systemfonts."),
   row.names=FALSE)
kdf <- kdf |>
   kableExtra::column_spec(2, bold=TRUE,
      extra_css="white-space: nowrap;") |>
   kableExtra::kable_styling(latex_options="scale_down")
kdf <- epub_kdf(kdf)
kdf
```

Note that it uses `grepl()` on the column `'family'` to subset the
set of all available fonts. To list all fonts, use: `subset_systemfonts()`.

#### Font family visualization

The results can be plotted by adding `do_plot=TRUE`, as a
basic way of viewing the resulting fonts. Note that there are no provisions
to prevent overlapping labels, so this approach is considered as a
"quick check" to view only a few fonts.

(ref:systemfonts-2) Graphical output as a "quick check" of fonts where that match the `subset_systemfonts()` output. These fonts show fonts where 'family' contains 'Helvetica'.

```{r systemfonts-2, fig.cap="(ref:systemfonts-2)", fig.alt="(ref:systemfonts-2)"}
ss <- subset_systemfonts(
   grepl("Helvetica", family),
   do_plot=TRUE)
```

#### Automatic font substitution

Another helpful feature of `systemfonts` is that it will substitute
a font, or a glyph within a font, when it cannot otherwise be found.
In Venndir, this approach helps ensure suitable Unicode characters
are used by default for example.

For example, the font family `'sans'` is replaced with the
appropriate corresponding font, which is important because it
differs on Windows, Mac, and linux architectures.
The typical defaults, for example:
'Arial' on Windows; 'Helvetica' on MacOS; 'DejaVu Sans' on Linux.

Similarly, when trying to use 'Arial' on a machine which lacks 'Arial',
`systemfonts` will substitute a suitable replacement,
for example 'DejaVu Sans' or 'Helvetica'.

The font substitutions can be pre-defined, see `systemfonts::font_fallback()`.

To review the font substitution, one can use `systemfonts::font_info()`
in the simple example below, shown in Table \@ref(tab:systemfonts-3).
There are many more options described in the systemfonts documentation.

```{r systemfonts-3-show, eval=FALSE}
data.frame(systemfonts::font_info("sans")[, 1:9])
```

```{r systemfonts-3, echo=FALSE}
idf <- data.frame(systemfonts::font_info("sans")[, 1:9]);
kdf <- knitr::kable(idf,
   caption=paste("Output from the font info function in systemfonts",
      "showing the font substitution for 'sans'."),
   row.names=FALSE)
kdf <- kdf |>
   kableExtra::column_spec(3, bold=TRUE,
      extra_css="white-space: nowrap;") |>
   kableExtra::kable_styling(latex_options="scale_down")
kdf <- epub_kdf(kdf)
kdf

```


## Venn Legends

One of many lessons learned from creating hundreds of Venn diagrams:
It is helpful to answer the most common questions upfront.
One of the most frequent questions is "How many items are in A?"
A straightforward answer can be provided in a table legend.

By default `venndir()` also calls `venndir_legender()` which adds a
legend to each figure.
This behavior can be skipped by using `draw_legend=FALSE`.

```{r legender-1, fig.height=5, fig.width=7, fig.cap="Default Venn diagram, showing a table legend in the bottom right corner.", fig.alt="Default Venn diagram, showing a table legend in the bottom right corner."}
setlist <- make_venn_test()
v <- venndir(setlist)
```

The legend is simple, one column `'Set'` lists each set name,
and one column `'Size'` lists the size of each set.
The background color matches the Venn diagram, and the text
is adjusted light or dark to maximize contrast.

The bottom row also includes 'Total' with the total number
of unique items represented by the Venn diagram.

When the Venn diagram displays [signed data](#signed-data), the legend
will also include [signed counts](#signed-counts).

(ref:legender-2) Venn diagram showing a table legend in the bottom right corner, this time the column 'Size' also includes counts tabulated by up and down directionality.

```{r legender-2, fig.height=5.5, fig.width=7, fig.cap="(ref:legender-2)", fig.alt="(ref:legender-2)"}
setlist_signed <- make_venn_test(do_signed=TRUE)
v <- venndir(setlist_signed)
```

Similarly, when the Venn diagram displays percentage values,
a column is added `'Percent'`.

(ref:legender-3) Venn diagram showing a table legend in the bottom right corner, now with new column 'Percent' since the Venn diagram also includes the percentage with each count.

```{r legender-3, fig.height=5.5, fig.width=7, fig.cap="(ref:legender-3)", fig.alt="(ref:legender-3)"}
v <- venndir(setlist_signed, show_labels="Ncps")
```

### Hide percent or signed labels

When percentage or signed labels are displayed in the Venn diagram,
the default legend will also include these labels. However, for simplicity
it may be preferred to hide these details from the legend, in order
to display a cleaner legend.
Several optional arguments can be used to hide specific components
of the legend.

* `legend_signed=FALSE`
* `legend_percentage=FALSE`
* `legend_total=FALSE`

(ref:legender-hide) Venn diagram showing counts, percentage values, and signed counts. A Venn legend is shown which includes the 'Set' and 'Size' columns, with the 'Percentage' column not shown due to argument 'legend_percentage=FALSE'.

```{r legender-hide, fig.height=5.5, fig.width=7, fig.cap="(ref:legender-hide)", fig.alt="(ref:legender-hide)"}
v <- venndir(setlist_signed,
   show_labels="Ncps",
   legend_signed=FALSE,
   legend_percentage=FALSE)
```

### Custom legend labels

The legend labels may be customized in `venndir()` with the
argument `legend_labels`. This argument defines the custom label
in the `Venndir` object so it is used in subsequent plots.

(ref:legender-4) Venn diagram showing a table legend in the bottom right corner, with custom labels for each set, for example 'set_A' is renamed 'Example Set A' in the legend.

```{r legender-4, fig.height=5.5, fig.width=7, fig.cap="(ref:legender-4)", fig.alt="(ref:legender-4)"}
v <- venndir(setlist_signed,
   legend_labels=c("Example Set A",
      "Example Set B",
      "Example Set C"))
```

Note that `legend_labels` will only customize the labels  in the
table legend, and not on the Venn diagram itself.
Sometimes it is necessary to customize the Venn diagram labels,
with the argument `setlist_labels`.

A common reason to have distinct labels in the figure and legend
is to impose line breaks in the figure which are not necessary
in the table.
At other times, it makes sense to have one
detailed label and one simplified label.

(ref:legender-5) Venn diagram showing a table legend in the bottom right corner, with custom labels for each set. 'set_A' is renamed 'Example Set A' in the legend.

```{r legender-5, fig.height=6.5, fig.width=7, fig.cap="(ref:legender-5)", fig.alt="(ref:legender-5)"}
v <- venndir(setlist_signed,
   expand_fraction=c(0.2, 0, 0, 0),
   setlist_labels=c("Example\nSet A",
      "Example\nSet B",
      "Example\nSet C"),
   y_inset=grid::unit(1, "lines"),
   legend_labels=c("Example Set A",
      "Example Set B",
      "Example Set C"))
```

### Additional alias labels

The argument `alias` can be used to supply simple labels,
which then pushes `legend_labels` into a new column `'Label'`.
The result may provide a helpful technique to associate extra
details to the Venn diagram labels.

Note that `alias` must be named using the set names, matching
the values in `names(setlist)`.

(ref:legender-6) Venn diagram showing a table legend in the bottom right corner, with custom labels for each set. 'Set A' in the Venn diagram is indicated 'A' in the legend, with a new column 'Contrast' with descriptive information.

```{r legender-6, fig.height=5.5, fig.width=7, fig.cap="(ref:legender-6)", fig.alt="(ref:legender-6)"}
v <- venndir(setlist_signed,
   legend_headers=c(Set="Set",
      Size="Size", Percentage="Percentage",
      Sign="Sign", Label="Contrast"),
   alias=c(set_A="A",
      set_B="B",
      set_C="C"),
   setlist_labels=c("Set A",
      "Set B",
      "Set C"),
   legend_labels=c(set_A="Dex - control",
      set_B="PGH - control",
      set_C="E2 - control"))
```

Figure \@ref(fig:legender-6) also uses the optional argument
`legend_headers` to define custom headings in the legend,
this adding 'Contrast' as a new header.


### Separate Size and Sign

The default legend combines the set size together with [signed counts](#signed-counts)
when the `Venndir` data contains [signed setlist](#signed-setlist).
In other words,the label $`r "30 (\u2193 16, \u2191 19)"`$ is displayed
in one field.

Two optional arguments provide alternatives to the default:

* `combine_size=FALSE` The total size will not be combined with
[signed counts](#signed-counts).

(ref:legender-size) Venndir legend with separate columns for 'Size' and 'Sign'.

```{r legender-size, fig.height=5.5, fig.width=7, fig.cap="(ref:legender-size)", fig.alt="(ref:legender-size)"}
venndir(setlist_signed,
   combine_size=FALSE)
```

* `combine_signed=FALSE` The [signed counts](#signed-counts) will not
be combined and displayed in parentheses.

(ref:legender-signed) Venndir legend with one column 'Size', whose values not delineated by parentheses.

```{r legender-signed, fig.height=5.5, fig.width=7, fig.cap="(ref:legender-signed)", fig.alt="(ref:legender-signed)"}
venndir(setlist_signed,
   combine_signed=FALSE)
```

* `combine_size=FALSE, combine_signed=FALSE` The total size and
[signed counts](#signed-counts) will each appear in separate columns.

(ref:legender-size-signed) Venn diagram with legend that shows separate columns with 'Size', and two additional columns with signed counts which do not have a header label.

```{r legender-size-signed, fig.height=5.5, fig.width=7, fig.cap="(ref:legender-size-signed)", fig.alt="(ref:legender-size-signed)"}
venndir(setlist_signed,
   combine_signed=FALSE,
   combine_size=FALSE)
```

### Legend color style

The default legend uses the same color fill as the Venn diagram.
However, the style can be customized using argument
'legend_color_style' to control both the
fill and border colors.

**Fill color**

* `"fill"` uses the Venn `set_colors` after applying `poly_alpha`
* `"nofill"` uses no fill color, inheriting the plot background
* `"greyfill"` uses light grey fill color

**Border color**

* `"border"` uses the Venn `set_colors` with no alpha transparency
* `"noborder"` uses no border
* `"greyborder"` uses medium-dark grey border
* `"blackborder"` uses black border

For example, to use no color fill, and black border:  
`legend_color_style=c("nofill", "blackborder")`

(ref:legender-bw) Venn diagram with a legend drawn using black border, and no background color.

```{r legender-bw, fig.height=5.5, fig.width=7, fig.cap="(ref:legender-bw)", fig.alt="(ref:legender-bw)"}
v <- venndir(setlist_signed,
   legend_color_style=c("nofill", "blackborder"))
```

The header can be customized as well, using arguments:
`header_color` for the text color, `header_bg` for background fill,
and `header_border` for the header border.

### Legend position

The legend itself is positioned using argument `legend_x` when called
by `venndir()`, or `x` when called by `venndir_legender()`.
The position recognizes the following `character` terms:

* x-axis position: 'left', 'center', 'right'
* y-axis position: 'top', 'center', 'bottom'

The default position is `"bottomright"`

Some fine-tuning is available with two more arguments:

* `x_inset` - `grid::unit` object applied when x is 'right' or 'left',
intended to apply an inset distance from the edge of the figure.
The default is `grid::unit(2, "lines")` which moves the legend two
character lines from the edge of the figure.
* `y_inset` - `grid::unit` object applied when x is 'top' or 'bottom',
which is applied the same way as described for `x_inset`. The default
is 2 character lines.

(ref:legender-bottomleft) Venn diagram showing the legend positioned in the bottom left corner, using argument 'legend_x' inside the `venndir()` function.

```{r legender-bottomleft, fig.height=5, fig.width=7, fig.cap="(ref:legender-bottomleft)", fig.alt="(ref:legender-bottomleft)"}
v <- venndir(setlist_signed,
   legend_x=c("bottomleft"))
```

For example, to move the legend closer to the bottom edge,
and closer to the left edge, see the following example.
Note the function `grid::grid.rect()` is used here to draw
a draw box around the plot region.

(ref:legender-inset) Venn diagram with legend in the bottom left corner, adjusted to the bottom edge, and moved more to the left of the figure. The grey box indicates the Venn figure region, the black box indicates the overall figure.

```{r legender-inset, fig.height=5, fig.width=7, fig.cap="(ref:legender-inset)", fig.alt="(ref:legender-inset)"}
v <- venndir(setlist_signed,
   x_inset=grid::unit(0, "lines"),
   y_inset=grid::unit(0, "lines"),
   legend_x=c("bottomleft"))
grid::grid.rect(gp=grid::gpar(fill=NA, lwd=2, col="grey"),
   vp=attr(v, "viewport"))
grid::grid.rect(gp=grid::gpar(fill=NA, col="black"))
```

Note the x-axis position is not the exact left
edge of the plot itself.
The legend is drawn within the Venndir `grid` viewport,
defined with fixed aspect ratio units 'snpc' (see `grid::unit()`).
As a result the Venndir viewport is roughly square, and if the
graphics device is wider or taller than square, extra
whitespace is imposed. The argument `x_inset` accepts
negative values that can position the legend further left
when necessary.

An alternative is to call `venndir()` with `draw_legend=FALSE`,
then separately call `venndir_legender()` to create the legend.
This process draws the legend in the overall viewport.

In Figure \@ref(fig:legender-inset-vp), the legend is positioned on the
exact bottom-left edge of the plot, shown by the grey line
drawn with `grid::grid.rect()`.

(ref:legender-inset-vp) Venn diagram with legend in the bottom left corner, this time with the legend touching the exact bottom left edge of the overall figure. The grey box indicates the Venn figure region, the black box indicates the overall figure.

```{r legender-inset-vp, fig.height=5, fig.width=7, fig.cap="(ref:legender-inset-vp)", fig.alt="(ref:legender-inset-vp)"}
v <- venndir(setlist_signed,
   draw_legend=FALSE)
grid::grid.rect(gp=grid::gpar(fill=NA, lwd=2, col="grey"),
   vp=attr(v, "viewport"))
grid::grid.rect(gp=grid::gpar(fill=NA, col="black"))
vl <- venndir_legender(v,
   x_inset=grid::unit(0, "lines"),
   y_inset=grid::unit(0, "lines"),
   x=c("bottomleft"))
```

Altogether, the use of `venndir()` with argument `expand_fraction`
should provide detailed control over the position of the Venndir
diagram. When necessary, the legend can be drawn separately
with `venndir_legender()` to afford exact control over the
position of the legend.

In future the plot method used for Venndir may be updated
to use 'npc' coordinates instead of 'snpc', which may enable
the legend to be freely positioned in the encompassing viewport,
not the Venndir viewport.
For now, this update is lower on the Todo list.


### Legend font size

The legend font size can be adjusted with `legend_font_cex` when called
by `venndir()`, otherwise it uses `font_cex` when called using
`venndir_legender()`.

(ref:legender-larger) Venn diagram with legend drawn in the bottom left corner, using font size adjusted to 1.3 times the default font size with argument 'font_cex'.

```{r legender-larger, fig.height=5, fig.width=7, fig.cap="(ref:legender-larger)", fig.alt="(ref:legender-larger)"}
v <- venndir(setlist_signed,
   legend_font_cex=1.3,
   x_inset=grid::unit(-3, "lines"),
   y_inset=grid::unit(0, "lines"),
   legend_x=c("bottomleft"))
```


### Legend custom fontfamily

The legend fontfamily will use the `fontfamily` defined
in the `Venndir` object (for example `v@metadata$fontfamily`).
For most cases, supplying a custom `fontfamily` to `venndir()`
will suffice, since it will apply the same custom font to the figure
and to the legend.

It is possible to use a custom font specifically for the legend.

This process requires calling `venndir_legender()` independently,
therefore the legend should not be drawn by `venndir()`.  
The steps are described and demonstrated below:

1. Call `venndir()` with `draw_legend=FALSE`, assign output
to a variable, in this case `v`.
2. Call `venndir_legender()` using the `Venndir` object `v` and
custom `fontfamily`.

(ref:legender-fontfamily) Venn diagram with legend drawn in the bottom right, using sans font for the Venn diagram labels, and serif font for the legend.

```{r legender-fontfamily, fig.height=5.5, fig.width=7, fig.cap="(ref:legender-fontfamily)", fig.alt="(ref:legender-fontfamily)"}
v <- venndir(setlist_signed,
   fontfamily="sans",
   draw_legend=FALSE)
vl <- venndir_legender(v,
   x="bottomright",
   x_inset=grid::unit(0, "lines"),
   y_inset=grid::unit(0, "lines"),
   font_cex=1.3,
   fontfamily="serif")
```


## Figure Scale and Position

The Venndir figure is automatically sized to fit within the
graphics device, and will re-adjust even when the figure is resized
in RStudio or an R graphics window. However, it can be helpful
to customize the scale and position further.

Common reasons to adjust scale and position:

* The outside labels extend beyond the page, usually
when text labels contains multiple lines.
* The Venndir figure, or the outside labels overlap
the legend.
* The figure could be resized to fit specific output dimensions.

Venndir uses the Venn circles and visible labels to create a
square "bounding box". It is square to keep the aspect ratio 1:1,
so circles remain circular. If the graphics device is not square,
there will be blank space around the Venndir figure. Sometimes
it helps to "zoom" the Venndir figure to fill the remaining
space.

The argument `expand_fraction` adjusts the whitespace "buffer" around
the Venndir diagram.

* `expand_fraction` above `0` makes the whitespace larger, shrinking the Venn circles.
* `expand_fraction` below `0` makes the whitespace smaller, enlarging the Venn circles.
* The font sizes are not adjusted.

The `expand_fraction` accepts four values, applied in clockwise order
from the bottom:  
`'bottom'`, `'left'`, `'top'`, `'right'`.  
Values are recycled to length=4, so `expand_fraction=0.1` is applied
to all four sides.

When the legend is shown, the Venndir figure is shifted up `0.18`.

(ref:expand-fraction-1) Venndir figure showing different values for 'expand_fraction'.

```{r expand-fraction-1, fig.ncol=2, fig.height=5, fig.width=7, out.width="50%", fig.cap="(ref:expand-fraction-1)", fig.alt="(ref:expand-fraction-1)", fig.subcap=c("Default.", "Default without legend.", "Default with title.", "Without legend, zoomed 10 percent.", "2-set default.", "2-set, zoomed 20 percent.")}
setlist <- make_venn_test()
venndir(setlist)
venndir(setlist, draw_legend=FALSE)
venndir(setlist, main="Main title")
venndir(setlist, draw_legend=FALSE, expand_fraction=-0.1)
venndir(setlist, sets=1:2)
venndir(setlist, sets=1:2, expand_fraction=c(0.1, 0, 0, 0) - 0.1)
```

:::: {.tipbox data-latex=""}

**Tips:**

The `expand_fraction` argument rules:

* When user-defined values are provided, or if they are stored in
the metadata of the `Venndir` object, `expand_fraction` is used
without any adjustments.
* When `expand_fraction` is not defined, these default adjustments
are applied:
   1. `0.07` is added to the top margin for each line in the
   `main` title.
   2. If the legend is shown, using `draw_legend=TRUE`, and
   the legend position `legend_x` includes 'top' or 'bottom',
   `0.2` is added to the corresponding 'top' or 'bottom' margin.
   It will not adjust the 'left' or 'right' margin.
   3. The top and bottom margins are reduced slightly.

* The `expand_fraction` is stored in the returned `Venndir` object
metadata when the plot is prepared, for example:
   * calling `plot()` or `render_venndir()`, or
   * calling `venndir()` with the default `do_draw=TRUE`.

::::


### Text Venn

Another convenient format for Venndir output is text Venn, intended
for R console output. This output is convenient for analysis work
on remote servers where interactive graphics are not available or
not as convenient.

The core function `textvenn()` mimics the input and arguments
of `venndir()` with some key limitations:

1. The input can only contain up to three sets.
2. The display only supports Venn layout, not proportional.
3. Output is a `data.frame` and not `Venndir` object.

The output follows the typical arrangement of text labels for
a Venn diagram, without the encompassing circular lines.
Colors and directional arrows are represented if the console
text output support these features, otherwise `unicode=FALSE`
will display `'^'` for "up", and `'v'` for "down".

The basic 2-way `textvenn()` output follows:

```{r textvenn-1, eval=FALSE}
setlist <- make_venn_test(set_names=LETTERS[1:3])
textvenn(setlist, sets=c(1, 3))
```

```{r textvenn-1-latex, echo=FALSE, eval=knitr::is_latex_output(), comment='', results='asis'}
setlist <- make_venn_test(set_names=LETTERS[1:3])
#
tvt <- capture.output(textvenn(setlist, sets=c(1, 3), htmlOut=TRUE))
tv_text <- html_to_latex_color(tvt);
cat(tv_text, sep="\n")
```


```{r textvenn-1-html, echo=FALSE, eval=knitr::is_html_output(), results='asis'}
setlist <- make_venn_test(set_names=LETTERS[1:3])
textvenn(setlist, sets=c(1, 3), htmlOut=TRUE)
```

The argument `sets` is described in [Consistent Set Colors],
and is used to include a subset of sets while maintaining
consistent color assignment to each set.

The 3-way `textvenn()` output follows:

```{r textvenn-3, eval=FALSE}
textvenn(setlist)
```

```{r textvenn-3-html, echo=FALSE, eval=knitr::is_html_output(), results='asis'}
textvenn(setlist, htmlOut=TRUE)
```

```{r textvenn-3-latex, echo=FALSE, eval=knitr::is_latex_output(), comment='', results='asis'}
tvt <- capture.output(textvenn(setlist, htmlOut=TRUE))
#
tv_text <- html_to_latex_color(tvt);
cat(tv_text, sep="\n")
```

These examples show only the Venn overlaps without
directionality. However, if given a [signed setlist](#signed-setlist)
it will also include directional tabulations:

```{r textvenn-4, eval=FALSE}
setlist3 <- make_venn_test(set_names=LETTERS[1:3],
   do_signed=TRUE)
textvenn(setlist3)
```

```{r textvenn-4-html, echo=FALSE, eval=knitr::is_html_output(), results='asis'}
setlist3 <- make_venn_test(set_names=LETTERS[1:3],
   do_signed=TRUE)
textvenn(setlist3, htmlOut=TRUE)
```




```{r textvenn-4-latex, echo=FALSE, eval=knitr::is_latex_output(), comment='', results='asis'}
setlist3 <- make_venn_test(set_names=LETTERS[1:3],
   do_signed=TRUE)
#
tvt <- capture.output(textvenn(setlist3, htmlOut=TRUE))
tv_text <- html_to_latex_color(tvt);
cat(tv_text, sep="\n")
```

The `template` argument described in [Signed Label Placement] is
also recognized. For example, `template='tall'` may improve the
layout especially when using `overlap_type='each'`, an option
for tabulating directionality that is described in [Overlap Type].

```{r textvenn-5, eval=FALSE}
textvenn(setlist3,
   overlap_type="each",
   template="tall")
```

```{r textvenn-5-html, echo=FALSE, eval=knitr::is_html_output(), results='asis'}
textvenn(setlist3,
   overlap_type="each",
   htmlOut=TRUE,
   template="tall")
```

```{r textvenn-5-latex, echo=FALSE, eval=knitr::is_latex_output(), comment='', results='asis'}
tvt <- capture.output(textvenn(setlist3,
   overlap_type="each",
   htmlOut=TRUE,
   template="tall"))
#
tv_text <- html_to_latex_color(tvt);
cat(tv_text, sep="\n")

```

#### Text Venn Items

A useful byproduct of `textvenn()` is that it returns a `data.frame`,
invisibly, which means it can be assigned to a variable but otherwise
is not printed to the output device. The `data.frame` includes the
count summary for each overlap, in addition to the `'items'` in
each set.

It is straightforward to use `textvenn()` to show the number of shared
and unique items, then use the `'items'` column to answer the
*most common next question*: "What are those?"

The example below captures `textvenn()` output into a variable `tv`.

```{r text-venn-items-1}
tv <- textvenn(setlist, sets=c(1, 3), do_plot=FALSE)
```

The items can be printed or used for subsequent analysis:

```{r text-venn-items-2}
tv$items
```
