# What's New by Version

```{r setup-news, include=FALSE}
knitr::opts_chunk$set(
   # fig.retina = 2, out.width = '100%'
   cache=TRUE,
   fig.align="center",
   fig.width=8,
   fig.height=7
)
```

```{r news-notes, eval=FALSE, echo=FALSE}
# Consider presenting as a table with two columns?
# - purpose is to keep track of documentation changes, alongside
#   occassional venndir R package version when it changes over time.
#
# Venndir-book v1.0.1         | Venndir version 0.0.57.900
#
# Venndir book should have an
# - Edition: 1, 2, 3 - book editioning
#    - major rewrite - new major chapter
# - Version: .1, .2, .3
#    - new diagram style
#    - feature additions
# - Addendum? .1.1, .1.2, .1.3
#    - minor corrections, typos
#    - small clarifications

```


```{r news-gt, echo=FALSE, eval=FALSE}
# library(yaml)
library(tibble)

whatsnew <- yaml::read_yaml("whatsnew.yml")
whatsnew[[1]]
whatsnew[[2]]
changelog <- whatsnew

tbl <- purrr::map_dfr(changelog, function(x) {
  tibble(
    book_version = x$book_version,
    venndir_version = x$venndir_version,
    change = x$changes
  )
})

tbl %>%
  gt(groupname_col = "book_version") %>%
  tab_header(title = "Venndir Book Changelog") %>%
  cols_label(venndir_version = "Venndir Version", change = "Change")

```

### v0.3.0 | 13 July 2025

* Edits to Preface, Introduction.

### v0.2.0 | 06 July 2025 | venndir v0.0.58.900 {-}

New chapters and sections:

* Inserted Chapter \@ref(whats-new-by-version): [What's New by Version]
* Inserted Chapter \@ref(glossary): [Glossary]
* Section \@ref(text-venn): [Text Venn]
* Section \@ref(figure-scale-and-position): [Figure Scale and Position]
* Section \@ref(custom-sign-label-colors): [Custom Sign Label Colors]
* Section \@ref(seehawer-kmt2-gene-venns): [Seehawer Kmt2 Gene Venns]

### v0.1.0 | 26 June 2025 | venndir v0.0.56.900 {-}

* Initial release, six chapters:

   1. What is a Venn Diagram?
   2. Data Import
   3. Venndir Basics
   4. Deeper Customizations
   5. Venndir Gallery
   6. References


