
# JamPolygon geometry


> *"I always thought quicksand was going to be a much bigger problem than it turned out to be" -- John Mulaney*

At some point along the journey creating Venndir, it became apparent
that geometry was going to be much more of an issue than I thought.
The quicksand of Venn diagrams turned out to be the geometry.

In an effort to minimize R dependencies used by Venndir, so that
using Venndir would be lightweight and convenient, a simple
object was created to support the geometry: `JamPolygon`

`JamPolygon` is a simple S4 class intended to store geometric objects,
and to enable a few basic features, summarized below:

1. Store coordinates for a closed polygon in a reliable way.
   * Support the ability to describe a "hole" inside a polygon.

2. Allow multiple polygons to be described as "one feature."
   * Polygons should always be "disjoint". That is, no two solid pieces
   are permitted to overlap, they would be combined using [union](#union).

3. Enable visualization in some convenient way.
   * Maintain correct aspect ratio, so that circles remain circular.
   * Utilize `grid` graphics for compatibility with other plots.

4. Add two specialized borders: inner, outer.
   * Display two adjacent borders without overlapping.
   * `innerborder` - appears just inside the boundary
   * `outerborder` - appears just outside the boundary
   * `border` - appears on the boundary directly
   * To be fair, this is a pet project. Don't judge.

5. Expand or shrink polygons by fixed size.
   * Respect multiple pieces, and holes, during the process.
   * Most of these functions are performed by a fantastic C++ library
   provided through an R package **polyclip** [@R-polyclip].

6. Enable other geometric functions.
   * Calculate: solid area, nearest point.
   * Transform: Rotate, transpose, resize.
   * Overlap: union, intersect, subtract.
   * Detect: point-in-polygon, nearest/farthest point.
   * Orient: clockwise or counterclockwise, solid or hole.
   * Label: outside, inside, filled.

Many of the features were necessary to use R package **polyclip** [@R-polyclip]
together with grid path grobs. `JamPolygon` provides a protective barrier
from some very detailed, very nuanced exceptions to the standard processing.


## Structure

`JamPolygon` is an S4 R object with one slot 'polygons' that contains
a `data.frame`. The `data.frame` has the following essential columns:

```{r jp_columns_essential, out.width="70%", echo=FALSE}
jp_cols1 <- data.frame(
   "Essential columns"=c(
      "'name'",
      "'x', 'y'"
   ),
   "Class"=c(
      "character",
      "list"
   ),
   "Content"=c(
      paste0("Unique identifier assigned to each polygon. ",
         "When 'label' is not present, this column is used as the default label."),
      paste0("Numeric coordinates in form of a list, with x,and y ",
         "coordinates for each polygon entry. One polygon entry per row, ",
         "which may contain one or more disjoint polygons.")
   )
)
jp_cols1
```

The `data.frame` may recognize any columns relevant to its use,
however the following optional columns are recognized by `plot()`
and other related functions.

```{r jp_columns_optional, out.width="70%", echo=FALSE}
jp_cols2 <- data.frame(
   "Optional columns"=c(
      "'label'",
      "'label_x', 'label_y'",
      "'label_color'",
      "'family'",
      "'fontsize'",
      "'fill'",
      "'border'",
      "'border.lwd'",
      "'border.lty'",
      "'innerborder'",
      "'innerborder.lwd'",
      "'outerborder'",
      "'outerborder.lwd'"
   ),
   "Class"=c(
      "character",
      "numeric",
      "character",
      "character",
      "numeric",
      "character",
      "character",
      "numeric",
      "character",
      "character",
      "numeric",
      "character",
      "numeric"
   ),
   "Content"=c(
      "Label associated with each polygon, used by plot() by default.",
      paste0("x,y coordinates used to place the label, ",
         "or NULL will use labelr_JamPolygon() to define coordinates dynamically. ",
         "Only one x,y coordinate is permitted per polygon row."),
      "R color used for text color when rendering labels.",
      paste0("String used as 'fontfamily' when rendering labels. ",
         "Note that labels are drawn using grid::textGrob() and not marquee::marquee_grob()."),
      "Font size used when rendering labels.",
      "R color used to fill the solid portion of the polygon, or NA for no color fill.",
      "R color used as the direct on-edge border for each polygon.",
      "Border line width",
      "Border line type, compatible with 'lty'.",
      "R color used as the inner solid region border, or NA for no inner border.",
      "Inner border line width",
      "R color used as the outer non-solid region border, or NA for no outer border.",
      "Outer border line width"
   )
)
jp_cols2
```

In most cases, calling `plot()` with extra arguments in ellipses `'...'`
will override the corresponding column values in the `JamPolygon`
object, and can be used to provide those values when they do not
already exist.


## Rules

* Each polygon is encoded into one row of the `jp@polygons` slot.
* Within one "polygon" - there may be multiple polygon pieces.
* Multi-part polygons can be split into the component parts
using `split_JamPolygon()`.
* All polygon pieces are subject to the "even-odd" rule, which
therefore has other requirements:
   * This rule essentially means each successive overlapping region
   flips from solid to transparent, then to solid, et cetera.
      * A figure "8" encodes two solid regions.
      * A circle "O" itself is solid, and a circle "o" inside that circle
      would therefore represent a hole. Another solid small circle may
      be placed within the hole, and so on.

   * For the purpose of `JamPolygon`, all polygon pieces should already
   be "disjoint" with no overlapping regions except when necessary
   to define a "hole".
   * No two solid polygons should have any regions that overlap, otherwise
   it would create a hole.
   * JamPolygon classifies every independent polygon "part" as either
   completely 'solid' or completely 'hole'.
   Any weird polygons that could otherwise be used to
   create solids and holes should be "flattened" and split into the
   component solid and hole parts.
   * These details are generally well beyond the scope of what
   is used by Venndir, but may be useful nonetheless.

* JamPolygon objects can be combined using `rbind2()` which appends
rows together, while expanding optional columns where present.


## Assembling a basic JamPolygon

The following example contains two polygons: 'polygon1', and 'polygon2'.
The columns 
The 'polygon1' is composed of two separate

```{r basic_jp}
df <- data.frame(
   name=c(
      "polygon1",
      "polygon2"),
    x=I(list(
        list(c(1, 4, 4, 1),
             c(2, 3, 3, 2)),
        c(5, 6, 6, 5))),
    y=I(list(
        list(c(1, 1, 4, 4),
             c(2, 2, 3, 3)),
        c(1, 1, 2, 2))),
    fill=c(
       "gold",
       "firebrick"))
jpdf <- new("JamPolygon",
   polygons=df)
plot(jpdf)
```



